---
title: "CPUEL: emulation and validation"
subtitle: "Reconstructing CPUEL row-by-row from raw HH/HL; discrepancy anatomy across all 17 Survey-Quarters"
author: "Einar Hjörleifsson"
date: "June 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(obus)
source("compute_cpue_emulation.R")
res <- readRDS("cpuel_validation_results.rds")
```

This document asks: can we reconstruct CPUEL row-by-row from raw HH and HL tables?
`compute_cpue_emulation()` does this and is validated here across all 17 Survey-Quarter
combinations and all available years. The goal is to understand what CPUEL computes
internally and to attribute row-count and CPUE discrepancies to their specific causes.

---

# Emulating CPUEL from raw HH/HL

`compute_cpue_emulation.R` reconstructs the ICES CPUEL row structure from raw
HH and HL tables. It is primarily a forensic tool — for understanding what CPUEL
does internally and for diagnosing discrepancies.

## Function overview

| Parameter | Default | Purpose |
|-----------|---------|---------|
| `haul_key` | `"short"` | `"short"` matches CPUEL's output key `(Survey:Year:Quarter:Platform:Gear:HaulNumber)`; `"full"` uses unambiguous `.id` |
| `collapse_sex` | `TRUE` | `FALSE` retains sex from HL — used for EVHOE in this validation (CPUEL retains F/M/U for EVHOE; all other surveys have sex = NA) |
| `filter_species_validity` | `"1"` | Codes contributing to non-zero CPUE values; does NOT affect the zero-fill species pool |

The zero-fill species pool is always built from **all** HL records in valid hauls
for the Survey-Quarter-Year, regardless of SpeciesValidity code — matching CPUEL
behaviour. Only CPUE values (non-zero rows) are filtered to `SpeciesValidity == "1"`.

## Validation: all 17 Survey-Quarters, all years

### Summary by Survey-Quarter

```{r summary-table}
res |>
  mutate(
    catch_contrib = (emul_catch - cpuel_catch) / cpuel_rows * 100,
    zf_contrib    = zf_diff / cpuel_rows * 100
  ) |>
  group_by(Survey, Quarter) |>
  summarise(
    `N years`          = n(),
    `Row diff %`       = round(mean(row_diff_pct,    na.rm = TRUE), 1),
    `Catch contrib %`  = round(mean(catch_contrib,   na.rm = TRUE), 1),
    `ZF contrib %`     = round(mean(zf_contrib,      na.rm = TRUE), 1),
    `CPUE diff %`      = round(mean(n_hour_diff_pct, na.rm = TRUE), 1),
    `Aphia diff`       = round(mean(aphia_diff,      na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(Survey, Quarter) |>
  knitr::kable(
    caption = paste(
      "Mean row diff decomposed into catch and zero-fill (ZF) contributions,",
      "all as % of cpuel_rows (additive: catch + ZF = total row diff).",
      "CPUE diff % is independent (uses total n_hour as denominator).",
      "Positive = emulation exceeds CPUEL."
    )
  )
```

**Near-perfect** (mean row diff < 1 %, catch and ZF both < 1 %):
ROCKALL Q3, SCOROC Q3, NS-IBTS Q1, NS-IBTS Q4.

**Reliable — small catch diff, near-zero ZF diff**:
IE-IGFS Q4 (row diff 2.6 %, entirely catch), NS-IBTS Q2/Q3 (< 3 %).

**Moderate row diff, driven by ZF**:
BITS Q3 (9.8 % total: 1.1 % catch + 8.7 % ZF), NS-IBTS Q2 (2.8 % total: 1.0 % catch + 1.8 % ZF).

**Moderate row diff, driven by catch data changes**:
SWC-IBTS Q1 (6.3 % total: 4.5 % catch + 1.8 % ZF — CPUE diff 7.3 %).

**Large row diff — ZF methodology artefact, CPUE reliable**:
EVHOE Q4 (37 % total: 5 % catch + 33 % ZF, CPUE diff 3 %). The April 2016
CPUEL computation omitted zero-fill rows; the emulation always generates full
zero-fill (see anatomy section).

**Large row diff — genuine HL data changes**:
SWC-IBTS Q4 (40 % total: 29 % catch + 10 % ZF, CPUE diff 38 %),
SCOWCGFS Q1 (26 % total: 20 % catch + 6 % ZF, CPUE diff 19 %).

**Very large row diff — both causes combined**:
BITS Q2 (127 % total: 116 % catch + 11 % ZF — vintage effect on sparse early-years data),
SCOWCGFS Q4 (96 % total: 66 % catch + 30 % ZF — 2019 CPUEL held only 3 aphias, a preliminary submission).

### Temporal patterns

Row differences have two independent causes with opposite implications:

- **Catch row diff** (`catch_diff_pct`): driven by HL data changes since the CPUEL snapshot. A large catch diff means real abundance estimates have shifted.
- **Zero-fill row diff** (`zf_diff_pct`): driven by differences in the zero-fill species pool and methodology. Zero-fill rows have n_hour = 0 and do not enter CPUE estimates.

#### Total row diff and its catch / zero-fill decomposition

```{r row-diff-decomp, fig.width=12, fig.height=10}
# Express all three as % of cpuel_rows so catch + ZF = total (additive)
res |>
  mutate(
    sq             = paste0(Survey, " Q", Quarter),
    catch_contrib  = (emul_catch - cpuel_catch) / cpuel_rows * 100,
    zf_contrib     = zf_diff / cpuel_rows * 100
  ) |>
  select(sq, Year,
         `Total row diff`  = row_diff_pct,
         `Catch component` = catch_contrib,
         `ZF component`    = zf_contrib) |>
  pivot_longer(-c(sq, Year), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric,
    levels = c("Total row diff", "Catch component", "ZF component"))) |>
  ggplot(aes(x = Year, y = value, colour = metric)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(
    values = c("Total row diff"  = "grey40",
               "Catch component" = "steelblue",
               "ZF component"    = "firebrick")
  ) +
  facet_wrap(~sq, scales = "free", ncol = 4) +
  labs(
    title    = "Row diff decomposed into catch and zero-fill components",
    subtitle = "All three expressed as % of cpuel_rows — catch + ZF = total row diff.",
    x = NULL, y = "Diff vs CPUEL (% of CPUEL rows)", colour = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

#### CPUE diff vs catch row diff

CPUE diff and catch row diff should track each other: both reflect genuine
changes in the HL data. When they diverge, it signals subsampling or CPUE
calculation differences beyond simple row counts.

```{r cpue-vs-catch, fig.width=12, fig.height=10}
res |>
  filter(!is.na(n_hour_diff_pct), is.finite(n_hour_diff_pct)) |>
  mutate(sq = paste0(Survey, " Q", Quarter)) |>
  select(sq, Year,
         `Catch row diff %` = catch_diff_pct,
         `CPUE diff %`      = n_hour_diff_pct) |>
  pivot_longer(-c(sq, Year), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Year, y = value, colour = metric)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_colour_manual(
    values = c("Catch row diff %" = "steelblue", "CPUE diff %" = "darkorange")
  ) +
  facet_wrap(~sq, scales = "free", ncol = 4) +
  labs(
    title    = "Catch row diff vs CPUE diff: both reflect HL data changes",
    subtitle = "Close tracking = consistent methodology. Divergence = subsampling or calculation differences.",
    x = NULL, y = "Diff vs CPUEL (%)", colour = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

Key contrasts:

- **EVHOE Q4**: mean total row diff 37 %, mean catch diff ~3 %, mean CPUE diff ~3 %.
  The large row diff is entirely in the zero-fill component. EVHOE's CPUEL is stored
  in two independent batches per year (see anatomy section below): a main species
  batch and a supplemental Atlantic mackerel batch. The April 2016 computation
  (years 2005–2015) included no zero-fill rows at all; the August 2018 and later
  computations do. The emulation always generates full zero-fill, producing 70–90 %
  row diff for the 2005–2015 years but near-zero catch diff throughout.
- **SWC-IBTS Q4**: both catch diff and CPUE diff ≈ 40 % — HL data has genuinely
  changed since the 2018 CPUEL snapshot.
- **NS-IBTS Q4, ROCKALL, SCOROC**: all three components < 1 % — near-perfect.

#### Species (aphia) count difference over time

```{r aphia-diff-time, fig.width=12, fig.height=9}
res |>
  mutate(sq = paste0(Survey, " Q", Quarter)) |>
  ggplot(aes(x = Year, y = aphia_diff)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_line(colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = aphia_diff == 0), size = 1.5, show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick")) +
  facet_wrap(~sq, scales = "free", ncol = 4) +
  labs(
    title    = "Aphia count difference: emulation minus CPUEL",
    subtitle = "Blue = perfect match; red = mismatch.",
    x = NULL, y = "Aphia diff (emulation − CPUEL)"
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

#### BITS: the sign flip

BITS Q1 and Q4 show the clearest temporal pattern. Before ~2008 the emulation
*overshoots* CPUEL: early CPUEL snapshots were computed on sparser HL data.
After ~2010 the emulation *undershoots*: BITS expanded to multi-vessel operations
and short-key collapsing dominates. CPUE differences remain small throughout.

```{r bits-detail, fig.width=9, fig.height=5}
res |>
  filter(Survey == "BITS", Quarter %in% c(1, 4)) |>
  mutate(Quarter = paste0("BITS Q", Quarter)) |>
  select(Quarter, Year, `Row diff %` = row_diff_pct,
         `CPUE diff %` = n_hour_diff_pct, `Aphia diff` = aphia_diff) |>
  pivot_longer(-c(Quarter, Year), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Year, y = value)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  annotate("rect", xmin = 2008, xmax = 2011, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.15) +
  annotate("text", x = 2009.5, y = Inf,
           label = "Transition\nperiod", hjust = 0.5, vjust = 1.3,
           size = 3, colour = "orange4") +
  geom_line(colour = "steelblue", linewidth = 0.6) +
  geom_point(size = 1.5, colour = "steelblue") +
  facet_grid(metric ~ Quarter, scales = "free_y") +
  labs(
    title    = "BITS Q1 and Q4: temporal decomposition",
    subtitle = "Orange band = multi-vessel expansion period (~2008–2011)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

#### EVHOE Q4: two-batch CPUEL structure and its effect on zero-fill

EVHOE Q4 CPUEL is stored as **two independent computation batches per year**,
each with its own `DateofCalculation`:

1. **Main species batch** — the bulk of the data. The computation date determines
   whether zero-fill was applied:
   - April 2016 vintage (years 2005–2015, 2016): **zero zero-fill rows**.
   - August 2018 vintage (years 1997–2004, 2017): full zero-fill included.
   - 2019+ vintage (2018 onward): full zero-fill included.
2. **Supplemental Atlantic mackerel batch** (DateofCalculation = 20181205, all
   years): aphia 127023 (*Scomber scombrus*) computed separately with proper
   zero-fill, adding 50–84 zero-fill rows per year.

```{r evhoe-batches}
dr_con("CPUEL") |>
  filter(Survey == "EVHOE", Quarter == 4) |>
  group_by(Year, DateofCalculation) |>
  summarise(
    n_aphias = n_distinct(aphia),
    n_catch  = sum(length_mm > 0, na.rm = TRUE),
    n_zf     = sum(length_mm == 0, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  collect() |>
  # label the two batch types
  mutate(
    Batch = if_else(n_aphias == 1, "Mackerel supplement", "Main species batch")
  ) |>
  filter(Batch == "Main species batch") |>
  select(Year, DateofCalculation, `N aphias` = n_aphias,
         `Catch rows` = n_catch, `ZF rows` = n_zf) |>
  arrange(Year) |>
  knitr::kable(
    caption = paste(
      "EVHOE Q4 main batch per year.",
      "April 2016 vintage (2005–2015) has zero ZF rows;",
      "August 2018 and later vintages include full zero-fill."
    )
  )
```

The emulation always generates full zero-fill from the current HL species pool.
For years with the April 2016 CPUEL vintage, this creates a structural mismatch:
CPUEL has essentially no zero-fill rows while the emulation generates ~15,000–22,000.
That alone accounts for 70–90 % row diff in those years, with catch differences
remaining below 5 %.

```{r evhoe-year-decomp, fig.width=9, fig.height=4}
res |>
  filter(Survey == "EVHOE", Quarter == 4) |>
  mutate(
    catch_contrib = (emul_catch - cpuel_catch) / cpuel_rows * 100,
    zf_contrib    = zf_diff / cpuel_rows * 100
  ) |>
  select(Year, `Total row diff`  = row_diff_pct,
         `Catch component` = catch_contrib,
         `ZF component`    = zf_contrib) |>
  pivot_longer(-Year, names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric,
    levels = c("Total row diff", "Catch component", "ZF component"))) |>
  ggplot(aes(x = Year, y = value, colour = metric)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  annotate("rect", xmin = 2004.5, xmax = 2016.5, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.12) +
  annotate("text", x = 2010.5, y = Inf,
           label = "April 2016 vintage\n(no ZF in CPUEL)",
           hjust = 0.5, vjust = 1.3, size = 3, colour = "orange4") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_colour_manual(
    values = c("Total row diff"  = "grey40",
               "Catch component" = "steelblue",
               "ZF component"    = "firebrick")
  ) +
  labs(
    title    = "EVHOE Q4: row diff decomposed by year",
    subtitle = "Orange band = years computed April 2016 (no zero-fill in CPUEL). All as % of cpuel_rows.",
    x = NULL, y = "Diff vs CPUEL (% of CPUEL rows)", colour = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

**Key takeaway:** The large row diff for EVHOE is overwhelmingly a zero-fill
methodology artefact, not a CPUE error. Catch row diffs remain < 5 % throughout,
and CPUE diffs < 3 %. The April 2016 CPUEL computation simply omitted zero-fill
rows for the main species batch; later computations include them. The emulation
matches the 1997–2004 and 2018+ years well once zero-fill is accounted for.

### Per-year detail

```{r per-year}
res |>
  select(Survey, Quarter, Year, cpuel_rows, emul_rows, row_diff_pct,
         cpuel_aphias, emul_aphias, aphia_diff, n_hour_diff_pct) |>
  arrange(Survey, Quarter, Year) |>
  knitr::kable(
    caption   = "Per Survey-Quarter-Year results",
    digits    = 1,
    col.names = c("Survey", "Q", "Year", "CPUEL rows", "Emul rows",
                  "Row diff %", "CPUEL aphias", "Emul aphias",
                  "Aphia diff", "CPUE diff %")
  )
```

## Known discrepancy causes

### Cause 1 — Short-key haul collapsing (emulation < CPUEL)

Multi-vessel surveys assign HaulNumbers 1–*n* independently per vessel; two
physically distinct hauls can share the same `(Platform, Gear, HaulNumber)` key.
The emulation's `distinct()` collapses them, producing fewer zero-fill rows than
CPUEL, which computes internally at the full `.id` level before truncating the key.

```{r short-key-example}
hh <- dr_con("HH") |>
  filter(Survey == "NS-IBTS", Year == 2018, Quarter == 1, HaulValidity == "V") |>
  collect()

tibble(
  Metric = c("Total valid hauls (.id)",
             "Distinct (Platform, Gear, HaulNumber)",
             "Hauls collapsed by short key"),
  Value  = c(nrow(hh),
             nrow(distinct(hh, Platform, Gear, HaulNumber)),
             nrow(hh) - nrow(distinct(hh, Platform, Gear, HaulNumber)))
) |> knitr::kable(caption = "NS-IBTS Q1 2018: short-key collapsing example")
```

Effect: ~5–15 % row undercounting for affected Survey-Years. Use
`haul_key = "full"` to avoid this at the cost of CPUEL output-format
compatibility.

### Cause 2 — CPUEL data vintage (emulation > CPUEL)

Each CPUEL record carries `DateofCalculation`. When that date precedes the
current HH/HL data, the local HL contains species not present in the CPUEL
snapshot, producing more emulation rows.

```{r vintage}
dr_con("CPUEL") |>
  group_by(Survey) |>
  summarise(
    `Earliest calc`    = min(DateofCalculation),
    `Latest calc`      = max(DateofCalculation),
    `N distinct dates` = n_distinct(DateofCalculation),
    .groups = "drop"
  ) |>
  collect() |>
  arrange(Survey) |>
  knitr::kable(caption = "DateofCalculation range per survey (format YYYYMMDD)")
```

SWC-IBTS last computed 2018; SCOWCGFS Q4 2019 has only 3 aphias in CPUEL
(preliminary submission) while the emulation correctly finds 99 from current HL.

### Cause 3 — Presence-only species (fixed in current version)

Species recorded with `SpeciesValidity` ≠ `"1"` and `NumberAtLength = NA`
(presence-only notes) had previously been excluded from the zero-fill pool
via `filter(NumberAtLength != 0)` before the species list was built. Fix:
build the pool from ALL HL records for valid hauls before that filter.

```{r presence-only}
all_aphias <- dr_con("HL") |>
  filter(Survey == "NS-IBTS", Year == 2018, Quarter == 1) |>
  distinct(aphia, SpeciesValidity) |>
  collect()

v1_aphias  <- all_aphias |> filter(SpeciesValidity == "1") |> pull(aphia)
only_nonv1 <- setdiff(all_aphias$aphia, v1_aphias)

dr_con("species") |>
  filter(aphia %in% only_nonv1) |>
  select(aphia, latin, species) |>
  collect() |>
  left_join(
    all_aphias |>
      filter(aphia %in% only_nonv1) |>
      group_by(aphia) |>
      summarise(validity_codes = paste(sort(unique(SpeciesValidity)), collapse = ","),
                .groups = "drop"),
    by = "aphia"
  ) |>
  arrange(latin) |>
  knitr::kable(
    caption = "NS-IBTS Q1 2018: presence-only species (SpeciesValidity ≠ '1') now included in zero-fill"
  )
```

### Cause 4 — Residual aphia gaps (NS-IBTS Q3, small)

NS-IBTS Q3 shows a consistent 1–3 aphia gap (emulation < CPUEL). Cause not
fully resolved; magnitude is small and does not affect CPUE values for species
that are present.

## Function status

```{r status-table}
tribble(
  ~Aspect, ~Status, ~Notes,
  "HaulValidity filter",       "✅ Correct",   "Filters HH to HaulValidity == 'V'",
  "Length units",              "✅ Correct",   "Converts LengthClass to mm",
  "Sex/stage collapse",        "✅ Handled",  "collapse_sex=TRUE for 7 surveys (CPUEL sex=NA); collapse_sex=FALSE for EVHOE (CPUEL retains F/M/U) — validation uses survey-specific setting",
  "CPUE calculation",          "✅ Correct",   "n_haul and n_hour via dr_add_n_and_cpue()",
  "Zero-fill species pool",    "✅ Fixed",     "Includes presence-only (NA NumberAtLength) species",
  "Short-key haul collapsing", "⚠️ Known gap", "distinct() collapses ~5-15 % of hauls in multi-vessel surveys",
  "CPUEL vintage mismatch",    "⚠️ By design", "Emulation uses current HL; CPUEL snapshots vary 2013–2026",
  "NS-IBTS Q3 aphia gap",      "⚠️ Residual",  "1-3 aphias consistently missing; cause unresolved"
) |>
  knitr::kable(caption = "compute_cpue_emulation() function status")
```

