---
title: "DATRAS Data Issues and Quality Checks"
subtitle: "Known limitations and inconsistencies requiring ICES attention, plus reproducible QC procedures"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  docx:
    toc: true
    toc-depth: 3
    number-sections: true
    reference-doc: ../reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)
library(obus)
```

> **Purpose of this document.** This page documents two categories of content: (1) **Known Issues** — problems and conventions requiring awareness or upstream correction; and (2) **Quality Checks** — reproducible R code to detect and verify these issues in the data.

---

## Part I: Known Issues

Each issue is classified by **where the fix lies**:

| Class | Meaning | Fix owner |
|-------|---------|-----------|
| **Specification gap** | The format spec is underspecified, ambiguous, or internally inconsistent | ICES documentation |
| **Software/API bug** | Bug or design flaw in an ICES tool | ICES software team |
| **Submission error** | Values in the submitted data violate the format specification | Data providers / ICES Datacenter |
| **Survey convention** | Deliberate survey practice that differs from the general format; not an error | No fix — user awareness |

---

## Specification Gaps {#sec-spec-gaps}

The DATRAS format specification is underspecified, inconsistent, or allows submission practices it does not formally define. Fix lies with ICES documentation and format governance.

### Column naming convention: legacy vs. standard {#sec-naming-convention}

Two naming conventions coexist in DATRAS, reflecting a change introduced by ICES around 2020. This is not a data quality issue per se, but a **system inconsistency** that causes silent failures in downstream code.

| Convention | Source | Example fields |
|------------|--------|----------------|
| **Legacy** | `getDATRAS()` (XML API); pre-~2020 DATRAS downloads | `Ship`, `HaulNo`, `StNo`, `LngtCode`, `HLNoAtLngt`, `SubFactor`, `TotalNo`, `HaulVal`, `HaulDur` |
| **Standard** | Parquet files; IMBUS product; current ICES convention | `Platform`, `HaulNumber`, `StationName`, `LengthCode`, `NumberAtLength`, `SubsamplingFactor`, `TotalNumber`, `HaulValidity`, `HaulDuration` |

Code written against one naming scheme fails silently or with cryptic errors against the other. {obus} and the IMBUS parquet use the standard names throughout.

**See:** [D2.2 §Column naming convention](../deliverables/D2.2_data_formats_metadata.qmd#column-naming-convention)

---

### IMBUS interim canonical names: `sex` and `aphia` {#sec-interim-names}

The ICES upload schema contains **naming inconsistencies that must be resolved upstream**:

| IMBUS name | ICES "new" name(s) | Problem |
|------------|--------------------|---------|
| `sex` | `SpeciesSex` (HL), `IndividualSex` (CA) | **Same concept, different names** — the ICES upload field list gives the same vocabulary two different field names depending on table. |
| `aphia` | `Valid_Aphia` (HL, CA), `AphiaID` / `ValidAphiaID` (derived products) | **Not in formal upload schema** — the datacenter-resolved WoRMS identifier is named differently depending on product type. |

These are **interim** canonical names used by IMBUS pending upstream resolution by the ICES Datacenter. Users should be aware that different systems (raw download vs. parquet vs. derived products) use different names for the same concept.

**See:** [D2.2 §IMBUS interim canonical names](../deliverables/D2.2_data_formats_metadata.qmd#sec-interim-names)

---

### Cross-table semantic ambiguities {#sec-cross-table-ambiguities}

Several fields share names across tables (HH, HL, CA) but carry different meanings. Users must be aware of these semantic shifts:

| Field | HL meaning | CA meaning |
|-------|------------|------------|
| **`sex`** | Visual, population-level assessment (bulk catch) | Determined by dissection (individual) |
| **`NumberAtLength`** | Sub-sample count per length bin; multiply by `SubsamplingFactor` for population estimate | Count of aged fish at that length class; used as raising key for age-length keys (ALK) |
| **`LengthClass`** | Lower boundary of a length frequency bin | Measured length of an individual fish |

Additionally, **gear fields are duplicated** across HH, HL, and CA with identical values — this is a submission artefact. HH is authoritative; HL and CA copies are redundant after joining.

**See:** [D2.2 §Cross-table ambiguities](../deliverables/D2.2_data_formats_metadata.qmd#cross-table-ambiguities)

---

### `TotalNumber` is not consistently defined across surveys {#sec-totalnumber-reconciliation}

The reporting format states that `TotalNumber = SubsampledNumber × SubsamplingFactor`. However, empirical reconciliation reveals **the two are not always equal**, and the **discrepancy pattern is survey-specific**:

**Three patterns identified:**

1. **All-sex total repeated on each sex row (Can-Mar)** — `TotalNumber` records the total count across *all* sexes, but the same value is repeated on every sex-specific row. After collapsing across sexes, most Can-Mar groups reconcile.

2. **Sex-specific `TotalNumber` (BTS, NL-BSAS)** — `TotalNumber` is reported separately for each sex. Comparing against any single sex's estimate gives apparent mismatches.

3. **Rounding (DYFS)** — `TotalNumber` and estimated totals both large (~2,000) but differ by ~2%, consistent with rounding during data entry.

**Implication:** `TotalNumber` **should not be relied upon as a portable cross-check** without knowing the survey-specific convention. The safe formula — `n_haul = sum(NumberAtLength × SubsamplingFactor)` from length rows — is self-consistent and does not depend on `TotalNumber`.

**See:** [Quality Checks §TotalNumber reconciliation](#sec-totalnumber-qc) for the reconciliation check code.

---

### Presence-only HL records: `TotalNumber`, `NumberAtLength`, and `LengthClass` all NA {#sec-presence-only}

**Issue:** Some HL records contain only a species identifier (`aphia`) with all quantity fields — `TotalNumber`, `NumberAtLength`, and `LengthClass` — submitted as `NA`. These are *presence-only* records: the submitting institute noted that a species was caught but provided no quantitative information.

**Root cause:** The DATRAS reporting format does not explicitly prohibit submitting an HL row with all quantity fields absent, so some institutes use this to flag incidental species presence without recording counts or lengths. This is an **ambiguous submission convention** — the DATRAS exchange format specification should either formally disallow it or define it as a distinct record type.

**Scope (NS-IBTS 2023 Q1):** 9 hauls affected (3 FR and 1 SE with DataType C; 4 NO and 1 NL with DataType R). These hauls produced **79 NA `n_haul` rows** out of 51,924 total non-NA rows — roughly 0.15% of haul × species combinations.

**In `HL_standardised` `type == "haul"` rows:** Groups where all `TotalNumber` values are `NA` produce `n_haul = NA`. These rows are **neither zeros nor valid catches** — they signal a species was recorded as present but not counted. Users who filter `n_haul > 0` will silently drop them.

**In `HL_standardised` `type == "length"` rows (fixed):** The opening `filter(NumberAtLength != 0)` previously dropped presence-only rows silently (SQL evaluates `NULL != 0` as NULL, not TRUE), causing 72 haul × species combinations to receive spurious zero rows. **Fixed in {obus}.**

**Recommended treatment:**
- Do not treat `n_haul = NA` rows as zero-catch entries.
- For binary presence analyses, treat `n_haul = NA` as "detected but not quantified".
- For abundance analyses, exclude `n_haul = NA` rows (do not zero-fill them).

---

## Software & API Bugs {#sec-software-bugs}

Bugs or design flaws in ICES tools (`icesDatras`, derived products). Fix lies in ICES software. {obus} works around both issues listed here.

### `xsi:nil` column name artefact in XML API responses {#sec-xsi-nil}

**Issue:** When `icesDatras::getCPUEAge()` or `icesDatras::getIndices()` returns data via the XML API, columns whose values are absent or sparse arrive with `xsi:nil="true"` embedded in their column names:

```
"Age_6 xsi:nil=\"true\""
"Age_7 xsi:nil=\"true\""
...
"Age_15 xsi:nil=\"true\""
```

**Root cause:** The ICES XML API marks absent elements using the W3C XML Schema nil convention (`xsi:nil="true"` attribute). The `icesDatras` XML parser does not strip this attribute — it leaks from the XML element into the parsed R column name.

**Consequence:** Malformed column names cause three downstream problems:

1. **Silent column mismatch on `bind_rows()`** — when iterating over multiple survey/year/quarter/species combinations, some data frames may carry `Age_6` (column had data) while others carry `Age_6 xsi:nil="true"` (column had no data). `dplyr::bind_rows()` treats these as distinct columns, producing a wider-than-expected table with mostly `NA` values in the duplicated columns.

2. **Type coercion warnings** — numeric values stored under a malformed name cannot be matched to the column type specification, triggering `NAs introduced by coercion` warnings even when the underlying data is valid.

3. **Inaccessible by standard column reference** — `df$Age_6` or `select(Age_6)` will fail silently; the backtick-quoted form `` df$`Age_6 xsi:nil="true"` `` is required.

**Confirmed affected record types:** `CPUEA` (Age_6 – Age_15 in NS-IBTS), `IDX` (Age_7 – Age_15 in NS-IBTS), `LT` (e.g., `Tickler xsi:nil="true"`). The artefact may appear intermittently in any XML-backed record type depending on survey, year, and quarter.

**{obus} workaround:** The internal `.strip_xsi_nil()` helper is applied to every data frame before `bind_rows()`:

```r
.strip_xsi_nil <- function(d) {
  names(d) <- sub(' xsi:nil="true"', "", names(d), fixed = TRUE)
  d
}
```

**See:** [Quality Checks §`xsi:nil` artefact](#sec-xsi-nil-qc)

---

### CPUEL haul key is too short for unambiguous joins {#sec-cpuel-haul-key}

**Issue:** CPUEL records lack `Country` and `StationName` fields present in HH. The short key (Survey, Year, Quarter, Platform, Gear, HaulNumber) is **not guaranteed unique**. This creates **join ambiguity** when attempting to map CPUEL back to HH metadata — over 2,400 short keys are non-unique in the current data.

**Root cause:** A product design decision in the CPUEL output format. CPUEL computes correctly internally, but the truncated key it publishes is unreliable for downstream joins.

When joining CPUEL to HH, **always use the full 8-field `.id` key if possible**, or manually verify uniqueness for your survey/year subset. This is one of the reasons the IMBUS `HL_standardised` product is preferred over CPUEL.

---

## Submission Errors {#sec-submission-errors}

Values in the submitted data that violate the DATRAS format specification. Fix lies with data providers (resubmission) or the ICES Datacenter (at-ingestion correction).

### Size-stratified subsampling submitted as DataType R {#sec-datatype-r-stratified}

**Issue:** The DATRAS schema intends DataType `P` for size-stratified hauls, but a minority (95 groups) are submitted as DataType `R`, **predominantly Danish vessels in NS-IBTS and BITS from 2024 onward** (82 of 95 groups). This is a **submission error that should be corrected at source**.

The row-level formula `NumberAtLength × SubsamplingFactor` remains correct, but `TotalNumber` reflects the stratum total, not the species-haul total.

**See:** [D2.2 §The DataType field](../deliverables/D2.2_data_formats_metadata.qmd#sec-datatype)

---

### DataType C with `SubsamplingFactor ≠ 1` {#sec-datatype-c-invalid}

For DataType C, `SubsamplingFactor` **must always equal 1**. Records with missing or `SubsamplingFactor ≠ 1` are **submission errors** and should be flagged to the ICES Datacenter.

**Scope:** At least **4,093+ records** have invalid `SubsamplingFactor` values for DataType C. Combined with the ~1,856 NA values (see [§SubsamplingFactor = NA](#sec-subfactor-na)), this represents a meaningful data quality issue affecting DataType C records.

---

### `SubsamplingFactor = NA` {#sec-subfactor-na}

**0.986% of valid HL records** have missing `SubsamplingFactor`. This is a **significant data submission error** affecting multiple surveys and DataTypes. `SubsamplingFactor` is required for CPUE arithmetic and propagates as `NA` in computed `n_haul` and `n_hour`.

**Distribution by DataType:**
- DataType R: 138,572 records (majority)
- DataType C: 1,856 records
- DataType S: 14 records

The issue spans multiple surveys, not only NS-IBTS.

---

### ICES statistical rectangle boundary mismatches {#sec-statrec-mismatch}

`StatisticalRectangle` in HH should match the ICES rectangle derived from `ShootLatitude`/`ShootLongitude`. Empirically, **only 0.23% of records** (308 of 132,989 records with complete data, excluding Can-Mar) show mismatches. Most surveys report correct rectangles; **DYFS has the highest mismatch rate** (139 records).

Mismatches can indicate:
- Data entry error
- Boundary cases (hauls spanning rectangle boundaries; rare)
- Coordinate rounding or precision issues
- Minor survey-specific spatial conventions

**Note:** Can-Mar (Northwest Atlantic) doesn't report `StatisticalRectangle` at all — it's 100% NA. This is not a mismatch issue but a geographic exclusion (see [§Can-Mar outside ICES rectangle coverage](#sec-can-mar-boundaries)).

---

## Survey-Specific Conventions {#sec-survey-quirks}

Deliberate survey practices that differ from the general format or from other surveys. Not errors — but users must account for them.

These include the HL grouping key constancy findings: the sampling protocol fields (`SpeciesValidity`, `SubsamplingFactor`) are expected to be constant within a natural group (`.id × aphia × sex × DevelopmentStage × SpeciesCategory`), but several surveys have by-design exceptions.

For the `StandardSpeciesCode` and `BycatchSpeciesCode` protocol flags that govern which species are required to be recorded per haul, see [Sampling protocols](sampling_protocol.qmd).

### `SpeciesValidity` varies within groups (by design) {#sec-specval-variation}

`SpeciesValidity` varies within **2,786 of ~2.6 million total groups (0.1%)**, predominantly BTS (1,289) and Can-Mar (1,272). This is **intentional** — full length-frequency rows (`SpeciesValidity = 1`) coexist with supplementary rows (totals-only, presence/absence) within the same group.

Most common co-occurrences: `{1, 5}` (1,281 groups) and `{4, 7}` (580 groups).

---

### `SubsamplingFactor` varies within groups (size-stratified subsampling) {#sec-subfactor-variation}

`SubsamplingFactor` varies within **389 of ~2.1 million valid groups (0.02%)** due to **size-stratified subsampling**. Small and large fish are subsampled at different rates within the same species group.

**Pattern by DataType:**
- **P** (293 groups) — all GB-SCT; the intended encoding across NS-IBTS, SCOWCGFS, SCOROC
- **R** (95 groups) — predominantly DK from 2024 onward; **submission error** (should be P; see [§DataType R](#sec-datatype-r-stratified))
- **C** (1 group) — anomaly

**Key formula:** `n_haul = Σ(NumberAtLength × SubsamplingFactor)` applied **row by row** is correct for all DataTypes, including size-stratified cases.

---

### `TotalNumber` varies by stratum (consequence of size-stratification) {#sec-totalno-variation}

`TotalNumber` and `SpeciesCategoryWeight` vary by stratum in **104 of ~2.1 million valid groups (0.005%)** as a direct consequence of size-stratified subsampling. This is a **data structure consequence**, not an error.

---

### SWC-IBTS ended 2010; superseded by SCOWCGFS {#sec-swc-ibts}

SWC-IBTS (Southwest Coast IBTS) records end in 2010 and are superseded by SCOWCGFS (Southwest Coast Continental Shelf Groundfish Survey). **Account for this break when analyzing time series.**

---

### SWC-IBTS contains non-standard sex code "-" {#sec-swc-ibts-sex-code}

**Issue:** SWC-IBTS contains **44,295 rows with sex = "-"** (hyphen/dash). This is a non-standard sex code; the DATRAS specification defines sex as `M` (male), `F` (female), `U` (unsexed), or `NA` (not recorded).

**Root cause:** Unknown. The "-" code may represent:
- Missing/undocumented data (equivalent to NA)
- A survey-specific convention with undefined meaning
- A data submission artefact

**Recommended treatment:** Treat sex = "-" as equivalent to `NA` (missing/unknown). This aligns with other surveys that report sex as `NA` and allows aggregation across surveys.

**Scope:** 44,295 rows in SWC-IBTS; no other survey exhibits this code.

---

### Can-Mar uses `SpeciesValidity = "5"` on rows with actual length data {#sec-canmar-specval}

`SpeciesValidity = "5"` is defined as "presence/absence only" — the expectation is that
such rows carry no length or count information (`LengthClass = NA`, `NumberAtLength = NA`).
BTS follows this convention: all 9 of its `{1, 5}` co-occurrence rows have `LengthClass = NA`.

**Can-Mar does not follow this convention.** Of 2,805 `SpeciesValidity = "5"` rows in
Can-Mar `{1, 5}` groups, 2,088 (74%) have `LengthClass` and `NumberAtLength` filled in.
Inspection of 535 Can-Mar groups where both `"1"` and `"5"` rows carry length data shows:

| Pattern | Groups | Share |
|---------|:------:|:-----:|
| `"5"` LengthClasses entirely distinct from `"1"` (additive data) | 226 | 42% |
| `"5"` LengthClasses partially overlap with `"1"` | 204 | 38% |
| `"5"` LengthClasses fully duplicated in `"1"` | 105 | 20% |

The reason for this split is not documented in the DATRAS submission format. One possibility
is that Can-Mar uses `SpeciesValidity` to distinguish two sampling subcategories within a
haul (e.g., different gear components, sub-tows, or species-ID confidence levels). Another
is that the `"5"` rows are a systematic submission artefact.

**Practical implication:** filtering to `SpeciesValidity == "1"` in Can-Mar discards real
length data in at least 42% of affected groups. The correct treatment is **unresolved** and
requires clarification from the Canadian data provider.

**See also:** [DATRAS FAQ § SpeciesValidity co-occurrence](#sec-specval-faq)

---

### Can-Mar is outside ICES statistical rectangle coverage {#sec-can-mar-boundaries}

Can-Mar (Canadian Maritime) survey **does not populate the `StatisticalRectangle` field** — all 11,372 valid Can-Mar records have `NA` for this field. This is **correct and expected**: `StatisticalRectangle` is an ICES grid system covering the **Northeast Atlantic only** (see [ICES maps](https://www.ices.dk/data/maps/pages/ices-statistical-rectangles.aspx)). Can-Mar operates in the **Northwest Atlantic (Canadian waters)**, which falls outside the ICES rectangle system.

Can-Mar does report complete shooting coordinates (latitude/longitude at tow start) for all records (11,372/11,372) and haul coordinates for most records (6,537/11,372 = 57%). For spatial analysis, **use coordinates for Can-Mar instead of relying on statistical rectangles**.

---

## Product Coverage Limitations {#sec-product-limitations}

Limitations in the scope of ICES-derived products. These are not bugs — they reflect the resources required to validate and produce derived products for each survey.

### CPUEL available for only 8 of 29 surveys; CPUEA/IDX vary by expert group {#sec-product-availability}

Derived products (CPUEL, CPUEA, IDX) are **developed and verified in cooperation with ICES Expert Groups**. This is time-consuming; newer surveys may lack products.

**Current coverage:**
- CPUEL: 8 surveys (BITS, EVHOE, IE-IGFS, NS-IBTS, ROCKALL, SCOROC, SCOWCGFS, SWC-IBTS)
- CPUEA and IDX: Vary by expert group agreement

---

---

# Part II: Quality Checks — Reproducible Code {#sec-qc-checks}

The following sections provide reproducible R code to **detect, verify, and quantify** the data issues documented above. All code chunks execute against the parquet-backed `dr_con()` lazy connection.

> **Note:** The code below reflects empirical verification performed in June 2026. All findings reported in Part I are based on this code and have been verified against the live data. These code blocks can be run interactively to confirm findings in real time.

---

## TotalNumber Reconciliation {#sec-totalnumber-qc}

Check whether `TotalNumber` matches the estimated total from length rows. This is a complex check because `TotalNumber` is scoped inconsistently across surveys:

```{r}
#| label: totno-reconcile
group_key <- c(".id", "aphia", "SpeciesCategory")

# Collect all HL data
all_hl <- dr_con("HL") |> collect()

# Length-frequency rows: estimated haul total
lf <- all_hl |>
  filter(!is.na(LengthClass), !is.na(NumberAtLength)) |>
  mutate(SpeciesCategory = coalesce(SpeciesCategory, -1L)) |>
  group_by(across(all_of(group_key))) |>
  summarise(estimated_n = sum(NumberAtLength * SubsamplingFactor, na.rm = TRUE),
            .groups = "drop")

# Catch-totals rows: stated total (take max per group — see text for caveats)
tot <- all_hl |>
  filter(is.na(LengthClass), !is.na(TotalNumber)) |>
  mutate(SpeciesCategory = coalesce(SpeciesCategory, -1L)) |>
  group_by(across(all_of(group_key))) |>
  summarise(TotalNumber = max(TotalNumber), .groups = "drop")

reconcile <- tot |>
  inner_join(lf, by = group_key) |>
  mutate(diff     = TotalNumber - estimated_n,
         rel_diff = diff / TotalNumber)

# Survey-level summary
reconcile |>
  filter(!is.na(rel_diff)) |>
  mutate(
    Survey = sub(":.*", "", .id),
    bucket = cut(abs(rel_diff),
                 breaks = c(0, 0.001, 0.01, 0.1, 0.5, Inf),
                 right  = FALSE,
                 labels = c("exact", "<1%", "1–10%", "10–50%", ">50%"))
  ) |>
  count(Survey, bucket) |>
  tidyr::pivot_wider(names_from = bucket, values_from = n, values_fill = 0L) |>
  select(Survey, exact, `<1%`, `1–10%`, `10–50%`, `>50%`) |>
  arrange(desc(exact + `<1%`)) |>
  knitr::kable(caption = "TotalNumber reconciliation by survey (survey-specific scoping creates apparent mismatches)")

# Show survey-specific patterns
cat("\nDiscordant groups by survey (>1% difference):\n")
reconcile |>
  filter(!is.na(rel_diff), abs(rel_diff) > 0.01) |>
  mutate(Survey = sub(":.*", "", .id)) |>
  group_by(Survey) |>
  summarise(
    n = n(),
    median_TotalNo = median(TotalNumber),
    median_est_n = median(estimated_n),
    median_rel_diff = round(median(rel_diff), 3),
    .groups = "drop"
  ) |>
  arrange(desc(n)) |>
  print()
```

---

## Size-Stratified Subsampling Submitted as DataType R {#sec-datatype-r-qc}

Identify hauls with size-stratified structure but DataType `R` (should be `P`):

```{r}
#| label: datatype-r-violations
sf_violations_R <- dr_con("HL") |>
  filter(SpeciesValidity == "1") |>
  group_by(.id, aphia, sex, DevelopmentStage, SpeciesCategory) |>
  summarise(n_sf = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_sf > 1) |>
  left_join(dr_con("HH") |> distinct(.id, DataType, Country), by = ".id") |>
  filter(DataType == "R") |>
  collect()

sf_violations_R |>
  group_by(Country) |>
  count(sort = TRUE) |>
  knitr::kable(caption = "Size-stratified DataType R hauls by country (95 groups total)")
```

---

## DataType C with Invalid `SubsamplingFactor` {#sec-datatype-c-qc}

Check for DataType C records where `SubsamplingFactor` is missing or not equal to 1 (all are data submission errors):

```{r}
#| label: datatype-c-violations
# Collect all data to avoid connection issues
all_hl <- dr_con("HL") |> collect()
all_hh <- dr_con("HH") |>
  select(.id, DataType) |>
  collect()

hl_with_dt <- all_hl |> inner_join(all_hh, by = ".id")

# Check DataType C
dt_c_total <- hl_with_dt |> filter(DataType == "C") |> nrow()
dt_c_na <- hl_with_dt |> filter(DataType == "C", is.na(SubsamplingFactor)) |> nrow()
dt_c_ne1 <- hl_with_dt |> filter(DataType == "C", !is.na(SubsamplingFactor), SubsamplingFactor != 1) |> nrow()
dt_c_invalid <- dt_c_na + dt_c_ne1

cat(sprintf("DataType C records with invalid SubsamplingFactor:\n"))
cat(sprintf("  SubsamplingFactor = NA: %d\n", dt_c_na))
cat(sprintf("  SubsamplingFactor ≠ 1: %d\n", dt_c_ne1))
cat(sprintf("  Total invalid: %d / %d (%.2f%%)\n\n", dt_c_invalid, dt_c_total, dt_c_invalid/dt_c_total*100))

# Show distribution of actual values
cat("Breakdown of SubsamplingFactor values in DataType C:\n")
dt_c_breakdown <- hl_with_dt |>
  filter(DataType == "C") |>
  count(SubsamplingFactor, sort = TRUE)
print(dt_c_breakdown |> head(10))
```

---

## Presence-only HL records {#sec-presence-only-qc}

Identify hauls and species groups where all quantity fields are absent. These produce `n_haul = NA` in `HL_standardised` `type == "haul"` rows and should not be treated as zero catches.

```{r}
#| label: presence-only-qc
presence_only <- dr_con("HL") |>
  filter(is.na(TotalNumber), is.na(NumberAtLength), is.na(LengthClass)) |>
  left_join(dr_con("HH") |> select(.id, DataType),
            by = ".id") |>
  group_by(Survey, Year, Quarter, Country, DataType) |>
  summarise(
    n_hauls   = n_distinct(.id),
    n_species = n_distinct(aphia),
    .groups   = "drop"
  ) |>
  collect()

cat("Total Survey/Year/Quarter/Country/DataType combinations affected:\n")
print(presence_only)

cat(sprintf("\nTotal hauls affected: %d\n",   sum(presence_only$n_hauls))  )
cat(sprintf("Total species groups: %d\n",     sum(presence_only$n_species)))
```

---

## Missing `SubsamplingFactor` {#sec-subfactor-na-qc}

Identify and quantify records with `NA` SubsamplingFactor. Note: empirical checking reveals this affects **0.986% of valid HL records** across multiple surveys, not just NS-IBTS.

```{r}
#| label: subfactor-na-qc
# Collect metadata once to avoid connection issues
all_hl <- dr_con("HL") |> collect()
all_hh <- dr_con("HH") |>
  filter(HaulValidity == "V") |>
  select(.id, DataType) |>
  collect()

# Merge and check
hl_with_meta <- all_hl |> inner_join(all_hh, by = ".id")

na_by_dt <- hl_with_meta |>
  filter(is.na(SubsamplingFactor)) |>
  count(DataType, sort = TRUE)

na_total <- sum(na_by_dt$n)
pct_total <- na_total / nrow(hl_with_meta) * 100

cat(sprintf("Total SubsamplingFactor = NA: %d / %d (%.3f%%)\n\n", na_total, nrow(hl_with_meta), pct_total))
cat("By DataType:\n")
print(na_by_dt)

cat("\n\nTop 15 surveys affected:\n")
na_by_survey <- hl_with_meta |>
  filter(is.na(SubsamplingFactor)) |>
  count(Survey, sort = TRUE) |>
  head(15)
print(na_by_survey)
```

---

## StatisticalRectangle Coverage by Survey {#sec-statrec-coverage-qc}

Some surveys do not populate the `StatisticalRectangle` field. Check coverage:

```{r}
#| label: statrec-coverage
statrec_coverage <- dr_con("HH") |>
  filter(HaulValidity == "V") |>
  group_by(Survey) |>
  summarise(
    n_total = n(),
    n_missing = sum(is.na(StatisticalRectangle)),
    .groups = "drop"
  ) |>
  collect() |>
  mutate(pct_missing = round(n_missing / n_total * 100, 2)) |>
  arrange(desc(pct_missing)) |>
  filter(pct_missing > 0)

if (nrow(statrec_coverage) > 0) {
  statrec_coverage |>
    knitr::kable(caption = "Surveys with missing StatisticalRectangle data")
} else {
  cat("All surveys report StatisticalRectangle for valid hauls.\n")
}
```

---

## ICES Statistical Rectangle Boundary Mismatch {#sec-statrec-qc}

Check for mismatches between reported and derived statistical rectangles (excluding surveys that don't populate this field):

```{r}
#| label: statrec-check
#| eval: false
hh <- dr_con("HH") |>
  filter(HaulValidity == "V", Survey != "Can-Mar") |>
  distinct(.id, Survey, lon = ShootLongitude, lat = ShootLatitude, ir = StatisticalRectangle) |>
  collect() |>
  drop_na(lon, lat, ir)

mismatches <- hh |>
  mutate(ir_derived = tryCatch(geo::d2ir(lat, lon), error = function(e) NA)) |>
  filter(!is.na(ir_derived)) |>
  filter(ir != ir_derived)

cat(sprintf("Total valid records with complete data (excluding Can-Mar): %d\n", nrow(hh)))
cat(sprintf("Mismatches: %d (%.2f%%)\n", nrow(mismatches), nrow(mismatches)/nrow(hh)*100))

if (nrow(mismatches) > 0) {
  cat("\nMismatches by survey:\n")
  mismatch_summary <- mismatches |>
    group_by(Survey) |>
    count(sort = TRUE)
  print(mismatch_summary)

  cat("\nFirst 5 examples:\n")
  mismatches |> slice_head(n = 5) |>
    select(.id, ir, ir_derived) |>
    knitr::kable(caption = "Sample StatisticalRectangle mismatches (reported vs. derived)")
}
```

---

## SpeciesValidity Variation Within Groups {#sec-specval-qc}

Identify groups where `SpeciesValidity` varies (intentional coexistence of record types):

```{r}
#| label: specval-variation
specval_v <- dr_con("HL") |>
  group_by(.id, aphia, sex, DevelopmentStage, SpeciesCategory) |>
  summarise(n_vals = n_distinct(SpeciesValidity), .groups = "drop") |>
  filter(n_vals > 1) |>
  collect()

specval_v |>
  left_join(dr_con("HH") |> select(.id, Survey) |> collect(), by = ".id") |>
  count(Survey, sort = TRUE) |>
  knitr::kable(caption = "Groups with varying SpeciesValidity by survey (2,786 total; 0.1%)")
```

---

## SubsamplingFactor Variation Within Groups {#sec-subfactor-qc}

Identify groups where `SubsamplingFactor` varies (size-stratified subsampling):

```{r}
#| label: subfactor-variation
sf_v <- dr_con("HL") |>
  filter(SpeciesValidity == "1") |>
  group_by(.id, aphia, sex, DevelopmentStage, SpeciesCategory) |>
  summarise(n_vals = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_vals > 1) |>
  left_join(dr_con("HH") |> distinct(.id, DataType, Country), by = ".id") |>
  collect()

sf_v |>
  count(Country, DataType, sort = TRUE) |>
  knitr::kable(caption = "Groups with varying SubsamplingFactor by country and DataType (389 total; 0.02%)")
```

---

## CPUEL Haul Key Ambiguity {#sec-cpuel-qc}

Check how many short keys (without Country/StationName) match multiple full `.id` keys in HH:

```{r}
#| label: cpuel-haul-key
cpuel_haul <- dr_con("CPUEL") |>
  mutate(.id_short = paste(Survey, Year, Quarter,
            Platform, Gear, HaulNumber,
            sep = ":")) |>
  distinct(.id_short) |>
  collect()

hh <- dr_con("HH") |>
  mutate(.id_short = paste(Survey, Year, Quarter,
            Platform, Gear, HaulNumber,
            sep = ":")) |>
  collect()

dupes <- hh |>
  distinct(.id, .id_short) |>
  inner_join(cpuel_haul, by = ".id_short") |>
  group_by(.id_short) |>
  filter(n() > 1) |>
  ungroup()

cat("CPUEL haul keys with multiple matches in HH:", nrow(dupes %>% distinct(.id_short)) / nrow(cpuel_haul) * 100, "%\n")

if (nrow(dupes) > 0) {
  dupes |> slice_head(n = 10) |>
    select(.id_short, .id) |>
    knitr::kable(caption = "Sample of ambiguous short keys")
}
```

---

## `xsi:nil` column name artefact {#sec-xsi-nil-qc}

Fetch a raw `icesDatras::getCPUEAge()` response and inspect the malformed column names directly. The example uses NS-IBTS 2022 Q1 (all species), which reliably produces the artefact on Age_6 – Age_15.

```{r}
#| label: xsi-nil-raw-names
#| echo: true
#| eval: false

# Raw response — bypasses obus entirely
raw_cpuea <- icesDatras::getCPUEAge("NS-IBTS", 2022, 1)

# Columns carrying the xsi:nil artefact
cat("Columns affected by xsi:nil artefact:\n")
print(grep('xsi:nil', names(raw_cpuea), value = TRUE))

cat("\nColumn count:", ncol(raw_cpuea), "\n")
```

The artefact can cause a wider-than-expected table when `bind_rows()` is called across API responses with inconsistent sets of affected columns. Demonstrate by simulating two frames where one call has the artefact on `Age_6` and the other does not (as would happen if the same age class had data in one survey-quarter but not another):

```{r}
#| label: xsi-nil-bind-problem
#| echo: true
#| eval: false

# Simulate the real-world mix: one frame with the artefact, one without
frame_with    <- raw_cpuea                          # Age_6 xsi:nil="true" present
frame_without <- raw_cpuea                          # same structure ...
names(frame_without) <- sub(' xsi:nil="true"', "",  # ... but artefact stripped from Age_6 only
                            names(frame_without), fixed = TRUE)
# Put the artefact back on Age_7+ to show a partial-strip scenario
for (col in grep('xsi:nil', names(frame_with), value = TRUE)[-1]) {
  names(frame_without)[names(frame_without) == sub(' xsi:nil="true"', "", col, fixed = TRUE)] <- col
}

result <- dplyr::bind_rows(frame_with, frame_without)
cat("nrow frame_with:   ", nrow(frame_with), "\n")
cat("nrow frame_without:", nrow(frame_without), "\n")
cat("ncol each frame:   ", ncol(frame_with), "\n")
cat("ncol after bind:   ", ncol(result), "  <-- extra column created\n")
cat("Duplicate Age_6 columns in result:\n")
print(grep("Age_6", names(result), value = TRUE))
```

Verify that the obus fix (`sub()` on `names()`) produces clean, accessible column names:

```{r}
#| label: xsi-nil-fix
#| echo: true
#| eval: false

clean <- raw_cpuea
names(clean) <- sub(' xsi:nil="true"', "", names(clean), fixed = TRUE)

cat("Artefact columns remaining after fix:", sum(grepl("xsi:nil", names(clean))), "\n")
cat("Age_6 now accessible via clean$Age_6:", !is.null(clean$Age_6), "\n")
cat("head(clean$Age_6):", head(clean$Age_6), "\n")
```

---

## Version history

| Date | Changes |
|------|---------|
| June 24, 2026 | Part I restructured into four issue classes (Submission Errors, Specification Gaps, Software & API Bugs, Survey-Specific Conventions). CPUEL haul-key section moved from Product Limitations to Software & API Bugs. HL grouping-key constancy violations merged into Survey-Specific Conventions. |
| June 23, 2026 | Added presence-only HL records issue and QC check. Added `xsi:nil` column name artefact issue and QC section. |
| June 22, 2026 | Reorganized into two parts: (1) Known Issues; (2) Quality Checks. Added cross-references to D2.2. |
| June 2026 | Consolidated from `DATRAS_data_issues.qmd` and `miscellaneous.qmd`. |
