---
title: "DATRAS Exchange Format: Reporting Format"
subtitle: "Field definitions, mandatory status, and survey-specific constraints across all 34 surveys"
date: "June 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

```{r setup}
#| include: false
library(httr)
library(rvest)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(DT)
library(stringr)
```

This page documents the DATRAS exchange format as reported by the [DATRAS Reporting
Format](https://datras.ices.dk/Data_products/ReportingFormat.aspx) web page. It covers all 34
surveys and five record types (HH, HL, CA, LT, FA), and answers two questions:

1. **Which fields are mandatory vs. optional?** — and does this vary across surveys?
2. **What are the allowed value ranges (Min / Max)?** — and how do these vary across surveys?

The data are scraped directly from the DATRAS web page and are current as of the document
build date.

---

## Scraping the reporting format {#sec-scrape}

```{r scrape}
#| cache: true
#| echo: true

DATRAS_URL <- "https://datras.ices.dk/Data_products/ReportingFormat.aspx"

get_hidden_fields <- function(page) {
  hid <- page |> html_elements("input[type='hidden']")
  setNames(html_attr(hid, "value"), html_attr(hid, "name"))
}

post_form <- function(handle, hidden, survey_val, record_val) {
  POST(
    handle = handle,
    path   = "/Data_products/ReportingFormat.aspx",
    encode = "form",
    body   = as.list(c(
      hidden,
      `__EVENTTARGET`   = "",
      `__EVENTARGUMENT` = "",
      `ctl00$ContentPlaceHolder1$ddl_Datatype_new` = survey_val,
      `ctl00$ContentPlaceHolder1$ddl_RecordTypes`  = record_val,
      `ctl00$ContentPlaceHolder1$BtnDisplay`       = "Display"
    ))
  ) |> (\(r) read_html(content(r, "raw")))()
}

extract_field_table <- function(page) {
  tbls <- page |> html_elements("table")
  result <- NULL
  for (tbl in tbls) {
    candidate <- tryCatch(html_table(tbl), error = \(e) NULL)
    if (!is.null(candidate) &&
        nrow(candidate) > 3 &&
        "FieldName" %in% names(candidate) &&
        "Mandatory" %in% names(candidate)) {
      result <- candidate
    }
  }
  result
}

# --- Fetch initial page ---
h  <- handle("https://datras.ices.dk")
r0 <- GET(handle = h, path = "/Data_products/ReportingFormat.aspx")
p0 <- read_html(content(r0, "raw"))
hidden0 <- get_hidden_fields(p0)

survey_opts <- p0 |>
  html_elements("#ContentPlaceHolder1_ddl_Datatype_new option") |>
  (\(x) data.frame(
    survey_val  = html_attr(x, "value"),
    survey_name = html_text(x) |> trimws(),
    stringsAsFactors = FALSE
  ))() |>
  filter(survey_val != "none")

record_opts <- p0 |>
  html_elements("#ContentPlaceHolder1_ddl_RecordTypes option") |>
  (\(x) data.frame(
    record_val  = html_attr(x, "value"),
    record_name = html_text(x) |> trimws(),
    stringsAsFactors = FALSE
  ))()

# --- Scrape all combinations ---
scrape_one <- function(survey_val, survey_name, record_val, record_name) {
  Sys.sleep(0.3)
  page <- tryCatch(
    post_form(h, hidden0, survey_val, record_val),
    error = \(e) NULL
  )
  if (is.null(page)) return(NULL)
  tbl <- extract_field_table(page)
  if (is.null(tbl)) return(NULL)
  tbl |>
    select(Order, FieldName, Width, Format, Mandatory,
           any_of(c("Min", "Max"))) |>
    mutate(
      Survey     = survey_val,
      SurveyName = survey_name,
      RecordType = record_name,
      Mandatory  = Mandatory == "True",
      .before    = 1
    )
}

combos <- crossing(survey_opts, record_opts)
results <- pmap(
  list(combos$survey_val, combos$survey_name,
       combos$record_val, combos$record_name),
  scrape_one
)
rf <- bind_rows(results)
```

```{r derived}
# Identify fields whose Min/Max vary by survey (shown separately below)
varying_fields <- rf |>
  group_by(RecordType, FieldName) |>
  filter(n_distinct(Min, na.rm = TRUE) > 1 | n_distinct(Max, na.rm = TRUE) > 1) |>
  ungroup() |>
  distinct(RecordType, FieldName)

# Reference field table: one row per RecordType × FieldName.
# Ranges are uniform across surveys for all non-varying fields, so first() is safe.
# Varying fields keep NA for Min/Max (per-survey detail shown in the constraints section).
rf_ref <- rf |>
  group_by(RecordType, Order, FieldName, Width, Format, Mandatory) |>
  summarise(Min = first(Min), Max = first(Max), .groups = "drop") |>
  rows_update(
    varying_fields |> mutate(Min = NA_real_, Max = NA_real_),
    by = c("RecordType", "FieldName"),
    unmatched = "ignore"
  ) |>
  arrange(RecordType, Order)

# Survey-varying min/max: fields where Min or Max differs across surveys
rf_ranges <- rf |>
  filter(!is.na(Min) | !is.na(Max)) |>
  group_by(RecordType, FieldName) |>
  filter(n_distinct(Min, na.rm = TRUE) > 1 |
         n_distinct(Max, na.rm = TRUE) > 1) |>
  ungroup()

# Survey list (drop the NS-IBTS_UNIFtest entry for range plots)
surveys_clean <- rf |>
  distinct(Survey, SurveyName) |>
  filter(!str_detect(Survey, "UNIFtest"))
```

---

## Overview {#sec-overview}

```{r overview-table}
rf |>
  group_by(RecordType) |>
  summarise(
    Fields     = n_distinct(FieldName),
    Mandatory  = n_distinct(FieldName[Mandatory]),
    Optional   = n_distinct(FieldName[!Mandatory]),
    Surveys    = n_distinct(Survey),
    .groups = "drop"
  ) |>
  arrange(RecordType) |>
  knitr::kable(caption = "Field counts per record type (mandatory status is uniform across all surveys).")
```

```{r survey-table}
rf |>
  distinct(Survey, SurveyName) |>
  arrange(Survey) |>
  knitr::kable(col.names = c("Survey code", "Full name"),
               caption = paste0("All ", nrow(distinct(rf, Survey)),
                                " surveys in the reporting format."))
```

**Key findings from the scrape:**

- Mandatory status is **identical for every survey** within a record type — the DATRAS
  exchange format does not vary field-level obligations by survey.
- Min / Max value ranges are defined for fields across all record types (HH, HL, CA, FA)
  and are likewise **uniform across surveys** — with the exception of five HH fields:
  `BottomDepth`, `Distance`, `GroundRopeWeight`, `HaulDuration`, and `WarpLength`,
  whose allowed ranges differ by survey and are shown separately in @sec-constraints.

---

## Field definitions per record type {#sec-fields}

Each table shows the complete field list for that record type. Mandatory fields are
highlighted. Min / Max are the allowed value ranges; blank cells indicate no constraint is
specified on the page.

```{r field-table-fn}
make_field_table <- function(record_type) {
  df <- rf_ref |>
    filter(RecordType == record_type) |>
    select(Order, FieldName, Format, Width, Mandatory, Min, Max) |>
    mutate(Mandatory = ifelse(Mandatory, "M", "O"))

  datatable(
    df,
    rownames = FALSE,
    filter   = "none",
    class    = "compact",
    options  = list(
      paging         = FALSE,
      scrollY        = "500px",
      scrollCollapse = TRUE,
      dom            = "t"
    ),
    caption  = paste(record_type, "— field definitions")
  ) |>
    formatStyle(
      "Mandatory",
      backgroundColor = styleEqual(c("M", "O"), c("#d4edda", "#f8f9fa")),
      fontWeight      = styleEqual("M", "bold")
    )
}
```

::: {.panel-tabset}

### HH — Haul Header

```{r tbl-hh}
make_field_table("HH")
```

### HL — Haul Length

```{r tbl-hl}
make_field_table("HL")
```

### CA — Catch at Age

```{r tbl-ca}
make_field_table("CA")
```

### LT — Litter

```{r tbl-lt}
make_field_table("LT")
```

### FA — Fish Activity

```{r tbl-fa}
make_field_table("FA")
```

:::

---

## Survey-specific field constraints {#sec-constraints}

All survey-specific variation in the reporting format is confined to allowed value ranges
(Min / Max) for five HH fields. The plots below show each survey's permitted range as a
horizontal segment. Surveys are sorted by their minimum value within each field.

```{r range-plots}
#| fig-height: 10
#| fig-width: 9

field_labels <- c(
  BottomDepth      = "Bottom Depth (m)",
  Distance         = "Distance (m)",
  GroundRopeWeight = "Ground Rope Weight (kg)",
  HaulDuration     = "Haul Duration (min)",
  WarpLength       = "Warp Length (m)"
)

rf_ranges |>
  filter(!str_detect(Survey, "UNIFtest")) |>
  mutate(FieldLabel = recode(FieldName, !!!field_labels)) |>
  group_by(FieldName) |>
  mutate(Survey = reorder(Survey, Min, FUN = \(x) mean(x, na.rm = TRUE))) |>
  ungroup() |>
  ggplot(aes(y = Survey)) +
  geom_segment(aes(x = Min, xend = Max, yend = Survey),
               linewidth = 1.2, colour = "#2166ac") +
  geom_point(aes(x = Min), size = 1.5, colour = "#2166ac") +
  geom_point(aes(x = Max), size = 1.5, colour = "#2166ac") +
  facet_wrap(~ FieldLabel, scales = "free_x", ncol = 2) +
  labs(x = "Allowed value range", y = NULL) +
  theme_bw(base_size = 9) +
  theme(panel.grid.minor  = element_blank(),
        axis.text.y       = element_text(size = 7))
```

### Full range data {#sec-range-table}

```{r range-table}
rf_ranges |>
  filter(!str_detect(Survey, "UNIFtest")) |>
  select(RecordType, FieldName, Survey, SurveyName, Mandatory, Min, Max) |>
  arrange(FieldName, Min) |>
  datatable(
    rownames = FALSE,
    filter   = "none",
    class    = "compact",
    options  = list(
      paging         = FALSE,
      scrollY        = "500px",
      scrollCollapse = TRUE,
      dom            = "t"
    ),
    caption  = "Survey-specific Min / Max constraints (HH fields only)."
  ) |>
  formatStyle(
    "Mandatory",
    backgroundColor = styleEqual(c(TRUE, FALSE), c("#d4edda", "#f8f9fa"))
  )
```
