---
title: "Sex determination in DATRAS HL data"
author: "Einar Hjörleifsson"
date: "June 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
  docx:
    toc: true
    toc-depth: 3
    number-sections: true
    reference-doc: ../reference.docx
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: load-packages
library(tidyverse)
library(obus)
library(santoku)
library(DT)
```

```{r}
#| label: load-data
hl      <- dr_con("HL")
species <- dr_lookup_species
```

# Overview of sex determination by survey

## Which surveys record any sex?

The table below shows, for each Survey × Quarter combination, which years contain at least one record with a real sex determination (`sex` is not `NA` and not `"U"`).

```{r}
#| label: survey-sex-overview
hl |>
  mutate(has.sex = ifelse(is.na(sex) | sex == "U", "no", "yes")) |>
  distinct(Survey, Quarter, Year, has.sex) |>
  collect() |>
  unite(col = "Survey", Survey, Quarter, sep = "_") |>
  arrange(Survey, Year) |>
  spread(has.sex, has.sex) |>
  DT::datatable()
```

## Proportion of species with sex determination over time

For each Survey × Quarter × Year, the proportion of species that have at least one individual with a real sex record is computed and displayed as a heatmap. A high proportion indicates that sex is systematically recorded for most species in that survey-year.

```{r}
#| label: compute-sex-proportion
d <-
  hl |>
  mutate(has.sex = ifelse(is.na(sex) | sex == "U", "no", "yes")) |>
  distinct(Survey, Quarter, Year, aphia, has.sex) |>
  group_by(Survey, Quarter, Year, has.sex) |>
  summarise(n_species = n_distinct(aphia), .groups = "drop") |>
  collect() |>
  unite(col = "Survey", Survey, Quarter, sep = "_") |>
  arrange(Survey, Year) |>
  tidyr::complete(Survey, Year, has.sex, fill = list(n_species = 0)) |>
  group_by(Survey, Year) |>
  mutate(p = n_species / sum(n_species)) |>
  ungroup()
```

```{r}
#| label: plot-sex-proportion
#| results: asis
#| fig-height: 1.5

cat("::: {.panel-tabset}\n\n")
for (sv in sort(unique(d$Survey))) {
  cat("##", sv, "\n\n")
  p <- d |>
    filter(Survey == sv, has.sex == "yes") |>
    mutate(p = santoku::chop(p, breaks = c(0, seq(0, 1, by = 0.25), 1))) |>
    ggplot(aes(Year, Survey, fill = p)) +
    geom_tile() +
    scale_fill_viridis_d(direction = -1) +
    labs(x = "Year", y = NULL, fill = "Proportion")
  print(p)
  cat("\n\n")
}
cat(":::\n")
```

# Species with real sex determination

This section focuses on which species have individual-level sex determined (i.e. `sex` is `"M"` or `"F"`), broken down by Survey × Quarter × Year × Country.

```{r}
#| label: compute-sex-species
sex_species <-
  hl |>
  filter(!is.na(sex), sex != "U") |>
  distinct(Survey, Quarter, Year, Country, aphia) |>
  collect() |>
  left_join(species |> collect(), by = c("aphia" = "aphia")) |>
  arrange(Survey, Quarter, Year, Country, latin)
```

## Species consistency across years

For each Survey × Quarter, the table below shows how consistently each species is recorded with real sex across years: `n_years` is the number of distinct years, `year_span` is the length of the observed period (last − first year + 1), and `consistency` is their ratio. A value of 1 means sex was recorded for every year in the species' observed window; values below 1 indicate gaps.

```{r}
#| label: species-consistency
sex_species_consistency <-
  sex_species |>
  unite(col = "SurveyQ", Survey, Quarter, sep = "_") |>
  group_by(SurveyQ, aphia, latin) |>
  summarise(
    n_years    = n_distinct(Year),
    year_first = min(Year),
    year_last  = max(Year),
    .groups    = "drop"
  ) |>
  mutate(
    year_span   = year_last - year_first + 1,
    consistency = n_years / year_span
  ) |>
  arrange(SurveyQ, desc(n_years))

DT::datatable(sex_species_consistency)
```

## Presence/absence heatmap across years

The heatmap below shows, for each Survey × Quarter, which species have sex determination recorded in each year. Each tile represents a species × year combination where at least one individual was sexed.

```{r}
#| label: prep-species-presence
sex_species_sq <- sex_species |>
  unite(col = "SurveyQ", Survey, Quarter, sep = "_")

# max species in any single survey tab, used to set fig height
fig_h_tab <- sex_species_sq |>
  group_by(SurveyQ) |>
  summarise(n = n_distinct(latin), .groups = "drop") |>
  pull(n) |>
  max() |>
  (\(x) ceiling(x * 0.25) + 2)()
```

```{r}
#| label: plot-species-presence
#| results: asis
#| fig-height: !expr fig_h_tab

cat("::: {.panel-tabset}\n\n")
for (sv in sort(unique(sex_species_sq$SurveyQ))) {
  cat("##", sv, "\n\n")
  p <- sex_species_sq |>
    filter(SurveyQ == sv) |>
    ggplot(aes(Year, reorder(latin, aphia))) +
    geom_tile() +
    labs(x = "Year", y = NULL) +
    theme(axis.text.y = element_text(size = 7))
  print(p)
  cat("\n\n")
}
cat(":::\n")
```

## Country coverage per species and survey

For each Survey × Quarter × species, the table below counts how many distinct countries contribute sex records. Species recorded by many countries are likely part of a standard sampling protocol; those recorded by only one country may reflect national practices.

```{r}
#| label: country-coverage
sex_species |>
  unite(col = "SurveyQ", Survey, Quarter, sep = "_") |>
  group_by(SurveyQ, aphia, latin) |>
  summarise(n_countries = n_distinct(Country), .groups = "drop") |>
  arrange(SurveyQ, desc(n_countries)) |>
  DT::datatable()
```

## Species with inconsistent sex recording

The table below lists species where sex recording is not consistent across years within a Survey × Quarter — i.e. where `consistency < 1` and the observed window spans more than one year. These species are candidates for further review to determine whether the gaps reflect biological absence, changing sampling protocols, or reporting inconsistencies.

```{r}
#| label: inconsistent-species
sex_species_consistency |>
  filter(year_span > 1, consistency < 1) |>
  arrange(SurveyQ, consistency, latin) |>
  select(SurveyQ, latin, aphia, n_years, year_first, year_last, year_span, consistency) |>
  DT::datatable()
```
