---
title: "D2.2 — Data Formats and Metadata"
subtitle: "IMBUS Deliverable D2.2 · Work Package 2"
author: "Einar Hjörleifsson (subcontracted to DTU Aqua)"
date: "June 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
  docx:
    reference-doc: ../reference.docx
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(obus)
library(dplyr)
library(knitr)
```

> **Project:** IMBUS — Implementing More and Better Use of ICES Survey Data  
> **Deliverable:** D2.2 — Data formats and meta-data  
> **Work Package:** WP2 — User requirements, data formats and meta-data  
> **Status:** DRAFT  
> **Due:** Month 11 (July 2026)

---

## Executive Summary

This document is the formal output of Task T2.3 (Data Formats and Meta-data) of the IMBUS
project. It defines the data format specification and metadata description for a standardised
representation of the ICES DATRAS bottom trawl survey exchange tables — the **HH** (haul
header), **HL** (haul-level catch at length), and **CA** (catch at age) record types.

The scope of D2.2 is the exchange data itself. Per the WP2-WP5 meeting of 1 June 2026, derived
products such as CPUE-by-age indices are outside WP2 scope. The CPUE-by-length calculation
pathway is described (see [CPUE Calculation Pathways]) as a transparent, reproducible procedure
that any user can apply to the standardised exchange tables, not as a deployed data service.

The principal design decisions are:

1. **Unified column naming.** A single, consistent set of column names (the "new" DATRAS variable names) is applied across all record types and surveys, regardless of API source or era.
2. **HL structure.** The HL table combines haul join keys, species identification, sampling
   protocol bookkeeping, catch totals, and length frequencies in a single flat structure. These
   are documented as distinct field groups. A proposed split into separate parquet files is
   pending final input from ICES and the IMBUS team (see [Proposed HL split design]).
3. **Parquet-based distribution.** HH, HL, and CA are stored as columnar parquet files enabling
   fast, column-selective querying without full downloads.
4. **Sentinel value removal.** The DATRAS convention of using −9 for missing values is replaced
   with proper `NA`.
5. **Species resolution via WoRMS AphiaID.** Historical species codes (ITIS TSN, NODC) are
   mapped to current WoRMS AphiaIDs for cross-era joins.
6. **Unique haul identifiers.** An 8-field composite key (`.id`) uniquely identifies each haul
   across all surveys, years, and countries.

The specification is implemented in the R package **{obus}**
(<https://github.com/einarhjorleifsson/obus>) and accompanied by illustrative code throughout.
Parquet files are currently hosted at <https://heima.hafro.is/~einarhj/datras/> (interim;
transfer to the ICES open access portal is planned under D2.4).

**Deliverable scope within this document.** D2.2 encompasses this document in full. §2
([ICES DATRAS Data Formats and Meta-data Descriptions]) documents the *current* DATRAS
exchange format — the primary reference for users working with raw data from any source.
§3 ([IMBUS Proposed Data Formats and D2.3 Product Specification]) specifies the IMBUS
standardised format and serves as the technical specification for the D2.3 data product
(parquet files). D2.3 and D2.4 are data deliverables (due M12); this document provides
their accompanying documentation.

---

## Background and Purpose {#sec-background}

The ICES DATRAS database holds fisheries-independent bottom trawl survey data submitted by
national institutes across the Northeast Atlantic, covering surveys from 1965 to the present.
The data are structured as three exchange tables — HH, HL, and CA — which together describe
each trawl haul, its catch by length class, and biological measurements on individual fish.

WP2 has four objectives:

1. Assess the user requirements of different user groups with respect to potential applications
   of survey data. *(Addressed in D2.1, April 2026.)*
2. Develop a data format that can be used to generate the products described in WP3–5.
3. Generate meta-data descriptions of the overall data.
4. Process the current data available in the DATRAS database into the new format, while
   correcting for any errors detected and taking account of zero observations and
   length/weight conversions.

Objective 1 has been delivered in D2.1 (IMBUS user requirements, April 2026, n = 55
respondents), which identified four dominant user challenges (Einar: Check if this is really correct)

* Complex table structure (HH/HL/CA hierarchy and join logic)
* Column name changes introduced around 2020 (legacy names vs. "new" names)
* Quality-control inconsistencies
* Slow downloads.

The current deliverable D2.2 addresses objectives 2 and 3 — the current and proposed data
formats and the associated meta-data descriptions.

### Scope

**In scope:**

- The three DATRAS exchange record types: **HH**, **HL**, and **CA**.
- Field-level metadata: names, types, units, descriptions, and known interpretation pitfalls.
- The field structure of HL and the proposed split design.
- The DataType field and its arithmetic implications for HL.
- Unique haul identification, species resolution, and length unit standardisation.
- The calculation pathway from raw exchange data to CPUE-by-length (as a documented procedure,
  not a deployed product).
- **Exploratory CPUE-by-age:** The calculation pathway from exchange data to age-disaggregated
  CPUE, using an age-length key (ALK) constructed from the CA table and applied to HL length
  frequencies. This is a transparent, user-driven procedure for research and exploration, not
  a formal assessment input.
- **Proposed intermediate (middle-data) parquet products:** Pre-processed derived tables that
  resolve DataType arithmetic, standardise length units, and separate the conceptual layers
  of HL — removing low-level implementation detail from the user's analytical path.
- Area standardisation: mapping haul positions to survey strata and ICES statistical areas.

**Out of scope:**

- **Official CPUEA and age-based survey indices:** The ICES Datacenter computes formal
  age-disaggregated CPUE products (CPUEA) and abundance indices (IDX) using survey- and
  species-specific age-slicing procedures agreed by ICES expert groups. Replicating or
  replacing these products is not part of WP2. They remain accessible via
  `dr_get("CPUEA", ...)` and `dr_get("IDX", ...)`.
- **Deployed data service:** WP2 produces a specification and supporting code. Deployment to
  the ICES open access portal is the subject of D2.4.
- **Per-survey metadata fact sheets:** Coverage, gear history, time-series breaks, and QC
  caveats at the survey level are addressed separately under user requirement UR-Q1.

---

## ICES DATRAS Data Formats and Meta-data Descriptions {#sec-datras}

The ICES DATRAS database distributes bottom trawl survey data as three exchange tables. Each
table is submitted independently per survey, year, quarter, and country; the tables are
linked by a common set of haul identification fields. The following sections describe the
current format of each table field by field, the metadata available for the surveys covered,
and known issues with the current formats.

The field specifications below are drawn from two machine-readable metadata sources
maintained in {obus}: `dr_lookup_fields` (field names, types, and descriptions — see
[Field and vocabulary reference]) and `dr_lookup_vocabulary` (controlled vocabularies for
categorical fields — see [Field and vocabulary reference]).

### Data Formats {#sec-formats}

#### HH — Haul Header {#sec-hh}

One row per haul. HH is the parent table; all child tables (HL, CA) join to it on `.id`.

##### Haul identification and join key

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `Survey` | `Survey` | chr | Survey acronym (e.g. `"NS-IBTS"`) |
| `Year` | `Year` | int | Year of survey |
| `Quarter` | `Quarter` | int | Quarter (1–4) |
| `Country` | `Country` | chr | ISO country code of contributing vessel |
| `Platform` | `Platform` | chr | Vessel identifier (SeaDataNet code) |
| `Gear` | `Gear` | chr | Gear code (e.g. `"GOV"`) |
| `StationName` | `StNo` | chr | Station identifier within survey |
| `HaulNumber` | `HaulNo` | int | Haul sequence number |
| `.id` | — | chr | Composite haul key (see [Unique Haul Identifier]) |

##### Haul outcome

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `HaulValidity` | `HaulVal` | chr | — | `"V"` = valid; `"I"` = invalid. Filter to `"V"` for standard analyses — [all codes](#vocab-haulval) |
| `DataType` | `DataType` | chr | — | Sampling protocol (see [Current Issues]) — [codes](#vocab-datatype) |
| `HaulDuration` | `HaulDur` | dbl | Minutes | Duration of tow |
| `ReasonHaulDisruption`^[Specified in ICES vocabulary but not in table]u ha | `ReasonHaulDisruption` | chr | — | Disruption code for invalid hauls |

##### Temporal

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `Month` | `Month` | int | — | Month of haul |
| `Day` | `Day` | int | — | Day of haul |
| `StartTime` | `TimeShot` | chr | HHMM GMT | Start time (`dr_add_starttime()` → POSIXct) |

`dr_add_date()` and `dr_add_starttime()` derive a tidy `date` (Date) and `time` (POSIXct)
column from these fields (and Year); the raw fields are then redundant.

##### Position and area

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `ShootLatitude` | `ShootLat` | dbl | Decimal degrees | Tow start latitude |
| `ShootLongitude` | `ShootLong` | dbl | Decimal degrees | Tow start longitude |
| `HaulLatitude` | `HaulLat` | dbl | Decimal degrees | Tow end latitude |
| `HaulLongitude` | `HaulLong` | dbl | Decimal degrees | Tow end longitude |
| `StatisticalRectangle` | `StatRec` | chr | — | ICES statistical rectangle; derivable from (start) position |
| `SurveyIndexArea`^[Specified in ICES vocabulary but not in table] | `SurveyIndexArea` | chr | — | Survey-specific stratum used in index calculations |
| `DepthStratum` | `DepthStratum` | chr | — | Survey-specific depth band |
| `BottomDepth` | `Depth` | int | m | Seabed depth at the haul location, measured from water surface |
| `DayNight` | `DayNight` | chr | — | `"D"` / `"N"` (15 min before/after sunrise/sunset) |

`StatisticalRectangle` and `DayNight` are derivable from position and time; they are retained
for convenience. `BottomDepth` is a property of the station location; it also determines
`WarpLength` in the gear setup.

##### Gear fields {#hh-gear}

Fields describe the tow and gear configuration. `SweepLength`, `DoorType`, and
`GearExceptions` are also present in HL and CA as a submission artefact; HH is the
authoritative source.

**Tow parameters** (how the haul was executed)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `Distance` | `Distance` | dbl | m | Distance between haul start and end |
| `SpeedGround` | `GroundSpeed` | dbl | knots | Ground speed of towing |
| `SpeedWater` | `SpeedWater` | dbl | knots | Trawl speed through water |
| `TowDirection` | `TowDir` | int | degrees | Towing direction; 360 = N |
| `MinTrawlDepth` | `MinTrawlDepth` | int | m | Shallowest point of trawl path (pelagic hauls) |
| `MaxTrawlDepth` | `MaxTrawlDepth` | int | m | Deepest point of trawl path (pelagic hauls) |

**Gear geometry** (sensor measurements taken during the haul)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `WingSpread` | `WingSpread` | dbl | m | Mean wing spread |
| `DoorSpread` | `DoorSpread` | dbl | m | Mean door spread |
| `NetOpening` | `Netopening` | dbl | m | Mean vertical net opening |

**Gear configuration** (rigging set before deployment)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SweepLength` | `SweepLngt` | int | m | Sweep length; survey- and quarter-specific recommended values |
| `DoorType` | `DoorType` | chr | — | Type of door used — [codes](#vocab-doortype) |
| `DoorSurface` | `DoorSurface` | dbl | m² | Door surface area |
| `DoorWeight` | `DoorWgt` | int | kg | Door weight |
| `WarpLength` | `Warplngt` | int | m | Warp length; determined by `BottomDepth` |
| `WarpDiameter` | `Warpdia` | int | mm | Warp diameter |
| `WarpDensity` | `WarpDen` | int | kg/m | Warp weight per linear metre |
| `Rigging` | `Rigging` | chr | — | Rigging method; survey-specific — [codes](#vocab-rigging) |
| `Tickler` | `Tickler` | int | — | Number of tickler chains |
| `Buoyancy` | `Buoyancy` | int | kg | Total buoyancy of net floats |
| `KiteArea` | `KiteDim` | dbl | m² | Kite area (pelagic hauls) |
| `GroundRopeWeight` | `WgtGroundRope` | int | kg | Total groundrope weight |
| `CodendMesh` | `CodendMesh` | int | mm | Codend mesh size |
| `PelagicSamplingType` | `PelSampType` | chr | — | Pelagic trawl sampling design — [codes](#vocab-pelsamtype) |
| `GearExceptions` | `GearEx` | chr | — | Deviations from standard gear; survey-specific — [codes](#vocab-gearex) |

##### Environmental fields {#hh-env}

Fields describe oceanographic and meteorological conditions at the time of the haul.
Direction fields use the convention 0 = no movement, 360 = flow/movement from south to
north, −1 = varying direction.

**Water column**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SurfaceTemperature` | `SurTemp` | dbl | °C | Surface water temperature |
| `BottomTemperature` | `BotTemp` | dbl | °C | Bottom water temperature |
| `SurfaceSalinity` | `SurSal` | dbl | PSU | Surface salinity |
| `BottomSalinity` | `BotSal` | dbl | PSU | Bottom salinity |
| `ThermoCline` | `ThermoCline` | chr | — | Thermocline observed? (`"Y"` / `"N"`); defined as ≥ 2 °C change over 10 m — [codes](#vocab-thermocline) |
| `ThermoClineDepth` | `ThClineDepth` | int | m | Depth of thermocline; populated only when `ThermoCline == "Y"` |
| `SecchiDepth` | `SecchiDepth` | dbl | m | Secchi disc depth (water clarity) |
| `Turbidity` | `Turbidity` | int | NTU | Turbidity |

**Currents and tides**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SurfaceCurrentDirection` | `SurCurDir` | int | degrees | Surface current direction |
| `SurfaceCurrentSpeed` | `SurCurSpeed` | dbl | m/s | Surface current speed |
| `BottomCurrentDirection` | `BotCurDir` | int | degrees | Bottom current direction |
| `BottomCurrentSpeed` | `BotCurSpeed` | dbl | m/s | Bottom current speed |
| `TidePhase` | `TidePhase` | int | min | Minutes before next high tide |
| `TideSpeed` | `TideSpeed` | dbl | m/s | Tidal speed |

**Wind and swell**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `WindDirection` | `WindDir` | int | degrees | Wind direction; calm = 0, varying = −1 |
| `WindSpeed` | `WindSpeed` | int | m/s | Wind speed |
| `SwellDirection` | `SwellDir` | int | degrees | Swell direction; varying = −1 |
| `SwellHeight` | `SwellHeight` | dbl | m | Swell/wave height |

---

#### HL — Haul-level Length {#sec-hl}

HL contains one row per length class per species per haul. It is the most complex of the three
exchange tables, combining haul join keys, species identification, sampling protocol
bookkeeping, catch totals, and length-frequency distributions in a single flat structure.
Understanding how these groups of fields interact is essential for correct CPUE arithmetic.

##### Haul identification and join key {#hl-haul-id}

HL shares the same eight haul-identification fields as HH. After joining HL to HH on `.id`,
these fields are redundant in HL and can be dropped.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `Survey` | `Survey` | chr | Survey acronym (e.g. `"NS-IBTS"`) |
| `Year` | `Year` | int | Year of survey |
| `Quarter` | `Quarter` | int | Quarter (1–4) |
| `Country` | `Country` | chr | ISO country code of contributing vessel |
| `Platform` | `Platform` | chr | Vessel identifier (SeaDataNet code) |
| `Gear` | `Gear` | chr | Gear code (e.g. `"GOV"`) |
| `StationName` | `StationName` | chr | Station identifier within survey |
| `HaulNumber` | `HaulNumber` | int | Haul sequence number |
| `.id` | `.id` | chr | Composite haul key (see [Unique Haul Identifier]) |

Three gear fields are duplicated from HH as a submission artefact (each exchange file is
self-describing). HH is the sole source of truth; the HL copies are redundant after joining
on `.id`.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `GearExceptions` | `GearExceptions` | chr | Non-standard gear modifications — [codes](#vocab-gearex) |
| `SweepLength` | `SweepLength` | dbl | Sweep length (metres) |
| `DoorType` | `DoorType` | chr | Door type code — [codes](#vocab-doortype) |

##### Species identification {#hl-species-id}

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `ValidAphiaID` | `Valid_Aphia` | int | Resolved WoRMS AphiaID; primary species join key across HL, CA, and external taxonomy tables |
| `SpeciesCode` | `SpeciesCode` | chr | Original species code as submitted (ITIS TSN, NODC, or WoRMS) |
| `SpeciesCodeType` | `SpeciesCodeType` | chr | Code system used: `"TSN"`, `"NODC"`, `"WoRMS"` |
| `DateofCalculation` | `DateofCalculation` | chr | Date on which the ICES datacenter resolved `SpeciesCode` to `ValidAphiaID` |

`SpeciesCode` and `SpeciesCodeType` reflect the original submission; `ValidAphiaID` is the
resolved, standardised identifier and should be used for all joins and aggregations.

##### Sampling protocol {#hl-sampling-protocol}

These fields describe how the catch was processed for a given species × sex × category group.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `SpeciesValidity` | `SpeciesValidity` | chr | Record type — governs which rows to include in products (see below) |
| `SpeciesCategory` | `SpeciesCategory` | chr | Subsampling category ID; up to 5 per species/sex/haul; always `"1"` for DataType C |
| `SubsamplingFactor` | `SubsamplingFactor` | dbl | Raise factor: e.g. 6 means 1 in 6 fish was measured |
| `SubsampledNumber` | `SubsampledNumber` | dbl | Count of fish actually measured in the subsample |
| `SubsampleWeight` | `SubsampleWeight` | dbl | Weight of the measured subsample (grams) |

> **Note (A2):** `SubsampledNumber` and `SubsampleWeight` are fields not confirmed in all
> surveys. Vaishav Soni and Adrianna Villamor have been asked to confirm their presence and
> exact semantics in the exchange data (action point A2, 1 June 2026).

**`SpeciesValidity`** controls what data are present in a row and which rows should be
included in derived products:

| Value | Meaning | Include in products |
|-------|---------|---------------------|
| `"1"` | Full record: lengths, numbers, weights | Yes |
| `"0"` | Invalid | No |
| `"4"` | Total number only; no lengths | Situation-dependent |
| `"5"` | Presence/absence only | Situation-dependent |
| `"6"` | Weight only | Situation-dependent |
| `"7"` | Numbers and weights; no lengths | Situation-dependent |
| `"10"` | Lengths and numbers; no weights | Situation-dependent |

Standard DATRAS products filter to `SpeciesValidity == "1"`. The {obus} functions expose this
via the `specval` argument (default `1L`).

**Constancy of sampling protocol fields.** These fields are in principle constant within a
natural group (`.id` × `ValidAphiaID` × `SpeciesSex` × `DevelopmentStage` × `SpeciesCategory`),
but empirical checking reveals two important exceptions:

- **`SpeciesValidity`** varies within 2,786 groups (predominantly BTS, 1,289 groups, and
  Can-Mar, 1,272 groups). The most common co-occurrences are `{1, 5}` (1,281 groups) and
  `{4, 7}` (580 groups), reflecting the coexistence of full length-frequency rows and
  supplementary rows (totals-only, presence/absence) within the same group. This is by design
  in those surveys.

- **`SubsamplingFactor`** varies within 389 groups due to **size-stratified subsampling** —
  small and large fish are subsampled at different rates within the same species group. This
  pattern occurs predominantly in GB-SCT vessels (293 groups, across NS-IBTS, SCOWCGFS,
  SCOROC) and DK vessels (82 groups, from 2024 onward in NS-IBTS and BITS). It occurs under
  both DataType P (293 groups, where it is the intended encoding) and DataType R (128 groups —
  see [Current Issues]).

  `TotalNumber` and `SpeciesCategoryWeight` also vary by stratum (~115 groups) as a direct
  consequence. The safe formula for the haul-level total is always
  `n_haul = sum(NumberAtLength × SubsamplingFactor)` applied at the individual row level —
  this is correct whether or not size-stratification is present.

```{r}
#| label: hl-constancy-checks
#| echo: true
#| eval: false

# Code used to derive the constancy statistics reported above.
# Uses old (parquet) column names: Valid_Aphia, IndividualSex, SubsamplingFactor.
library(obus)
library(dplyr)

hl <- dr_con("HL")

# Key grouping: natural species-category group within a haul
grp <- c(".id", "Valid_Aphia", "IndividualSex", "DevelopmentStage", "SpeciesCategory")

# --- SpeciesValidity: groups where it takes more than one distinct value ---
specval_violations <- hl |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(SpeciesValidity), .groups = "drop") |>
  filter(n_vals > 1)

# Total violating groups
specval_violations |> count()                        # 2,786

# By survey
specval_violations |>
  left_join(hl |> distinct(.id, Survey), by = ".id") |>
  count(Survey, sort = TRUE)                         # BTS 1,289; Can-Mar 1,272

# Most common co-occurrences of SpeciesValidity values
hl |>
  semi_join(specval_violations, by = grp) |>
  group_by(across(all_of(grp))) |>
  summarise(combo = paste(sort(unique(SpeciesValidity)), collapse = ","),
            .groups = "drop") |>
  count(combo, sort = TRUE)                          # {1,5}: 1,281; {4,7}: 580

# --- SubsamplingFactor: groups where it takes more than one distinct value ---
sf_violations <- hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_vals > 1)

# Total violating groups
sf_violations |> count()                             # 389

# By country and survey
sf_violations |>
  left_join(hl |> distinct(.id, Survey, Country), by = ".id") |>
  count(Country, Survey, sort = TRUE)                # GB-SCT 293; DK 82 (from 2024)

# By DataType
sf_violations |>
  left_join(dr_con("HH") |> distinct(.id, DataType), by = ".id") |>
  count(DataType)                                    # P: 293; R: 128

# --- TotalNumber: groups where it varies across strata ---
hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_vals > 1) |>
  left_join(hl |> distinct(.id, Survey), by = ".id") |>
  count(Survey, sort = TRUE)                         # ~115 groups total
```

##### Catch totals {#hl-catch-totals}

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `TotalNumber` | `TotalNumber` | dbl | — | Total catch per group; semantics depend on DataType (see [Current Issues]) |
| `SpeciesCategoryWeight` | `SpeciesCategoryWeight` | dbl | g | Total catch weight per group; also DataType-dependent |

`TotalNumber` and `SpeciesCategoryWeight` are repeated identically across every length row for
a given `.id` × `ValidAphiaID` × `SpeciesSex` × `DevelopmentStage` × `SpeciesCategory` group
in most records. **Do not sum `TotalNumber` across length rows** — a species with 30 length
bins produces 30 identical copies of the same total. When size-stratified subsampling is used,
`TotalNumber` differs between strata and reflects the stratum total, not the species-haul total.

**Worked example — size-stratified subsampling (NS-IBTS 2025, DK vessel, whiting
AphiaID 126437):**

| Stratum | Length range | `SubsamplingFactor` | Rows measured | `TotalNumber` (stratum) |
|---------|-------------|---------------------|---------------|------------------------|
| Small fish | 100–160 mm | 5.9 | 61 | ≈ 360 |
| Large fish | 210–420 mm | 1.0 | 168 | 168 |

Correct haul total: **360 + 168 = 528 fish**. Direct use of either stratum `TotalNumber` is
wrong.

The correct approach for all DataTypes R, S, and P:

```r
hl |>
  filter(SpeciesValidity == "1") |>
  left_join(hh |> select(.id, DataType, HaulDuration), by = ".id") |>
  group_by(.id, ValidAphiaID, SpeciesSex, DevelopmentStage, SpeciesCategory) |>
  summarise(n_haul = sum(NumberAtLength * SubsamplingFactor), .groups = "drop")
```

##### Length frequency {#hl-length-freq}

One row per length class within a species × sex × development stage × category group.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `LengthClass` | `LengthClass` | int | Length class value; unit determined by `LengthCode` |
| `LengthCode` | `LengthCode` | chr | Unit flag: `"mm"` = millimetres, `"cm"` = 0.5 cm classes, `"1"` = 1 cm classes |
| `LengthType` | `LengthType` | chr | Measurement convention: total length, fork length, mantle length, etc. |
| `NumberAtLength` | `NumberAtLength` | dbl | Count in the measured sub-sample at this length class; semantics depend on DataType |
| `SpeciesSex` | `IndividualSex` | chr | Population-level sex: `"M"` / `"F"` / `"U"` / `NA`; determined by visual inspection of the bulk catch |
| `DevelopmentStage` | `DevelopmentStage` | chr | Non-invasive maturity or developmental stage observation |

> **Note (A2):** `LengthType` is particularly relevant for species where the standard length
> convention differs by country or era. `DevelopmentStage` prevalence and the conditions under
> which it is populated require confirmation (action point A2, 1 June 2026).

`SpeciesSex` and `DevelopmentStage` are grouping dimensions within the length frequency —
they define distinct sub-populations measured separately. Both must be included in any
grouping key before aggregating across length classes.

**Naming note:** The parquet file retains the legacy name `IndividualSex` for the
population-level sex field in HL. This is a misnomer: in HL the sex is assessed visually on
the bulk catch, not by individual dissection. The proposed standard name `SpeciesSex` is used
throughout this document to distinguish it from `IndividualSex` in CA, which is
dissection-based.

##### Administrative {#hl-admin}

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `RecordHeader` | `RecordHeader` | chr | Record type identifier; always `"HL"` |

---

#### CA — Catch-at-age {#sec-ca}

CA contains biological measurements on fish selected for ageing. One row per measured
individual (most rows) or per length class for pooled area-level samples.

CA shares the same eight haul join fields as HH and HL. After joining on `.id`, the haul-level
fields are redundant.

##### CA-specific fields (post-join)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `ValidAphiaID` | `Valid_Aphia` | int | — | Resolved WoRMS AphiaID |
| `AreaCode` | `AreaCode` | chr | — | Aggregation area code; check `AreaType` before joining to HH |
| `AreaType` | `AreaType` | chr | — | `"H"` = haul-level (safe to join); other values = area-pooled |
| `FishID` | `FishID` | chr | — | Individual fish identifier within haul |
| `IndividualSex` | `IndividualSex` | chr | — | Sex by dissection: `"M"` / `"F"` / `"U"` — distinct from `SpeciesSex` in HL |
| `LengthClass` | `LengthClass` | int | see `LengthCode` | Measured length of individual fish |
| `LengthCode` | `LengthCode` | chr | — | Length unit flag |
| `LengthType` | `LengthType` | chr | — | Total length, fork length, mantle length, etc. |
| `IndividualAge` | `IndividualAge` | int | Years | Otolith-based age estimate |
| `AgePlusGroup` | `AgePlusGroup` | chr | — | `"+"` if fish is assigned to the plus group |
| `AgeSource` | `AgeSource` | chr | — | Tissue type (otolith, scale, etc.) |
| `AgePreparationMethod` | `AgePreparationMethod` | chr | — | Preparation method |
| `OtolithGrading` | `OtolithGrading` | chr | — | Reader quality score |
| `IndividualWeight` | `IndividualWeight` | dbl | g | Individual body weight |
| `LiverWeight` | `LiverWeight` | dbl | g | Liver weight when sampled |
| `IndividualMaturity` | `IndividualMaturity` | chr | — | Maturity stage code |
| `MaturityScale` | `MaturityScale` | chr | — | Maturity scale used (M6 or SMSF from 2021+) |
| `NumberAtLength` | `NumberAtLength` | dbl | — | Count at this length/age class (pooled rows only) |

> **Important:** When `AreaType != "H"`, age samples are pooled across multiple hauls within
> an area. Joining these rows to a specific HH record is ambiguous. Always check `AreaType`
> before assuming a 1:1 join to HH.

---

#### CPUE and Derived Products {#sec-cpue-products}

ICES produces several derived products from the exchange tables, accessible without
downloading raw data:

| Product | Access | Description |
|---------|--------|-------------|
| CPUE per length (CPUEL) | `dr_get("CPUEL", ...)` | Official ICES CPUE-by-length product |
| CPUE per age (CPUEA) | `dr_get("CPUEA", ...)` | Age-disaggregated CPUE |
| Survey indices (IDX) | `dr_get("IDX", ..., aphia = ...)` | Age-based survey indices |
| Catch weight (CW) | `dr_get("CW", ..., aphia = ...)` | Catch weight by species |
| Flex file (FL) | `dr_get("FL", ...)` | Survey-specific additional fields |
| Litter assessment (LT) | `dr_get("LT", ...)` | Marine litter per haul |

The official CPUEA and IDX products embed survey- and species-specific decisions (ALK
construction, plus-group handling, stratum weighting) agreed by ICES expert groups. These
decisions are not encoded in the exchange tables and cannot be replicated from exchange data
alone. WP2 does not aim to replicate these products; see [Standardized Code] for the IMBUS
exploratory CPUE-by-age approach.

---

### Meta-data Descriptions {#sec-metadata}

#### Surveys covered

The specification applies to all ICES surveys available in DATRAS with valid hauls
(`HaulValidity == "V"`), excluding test surveys. The table below is generated from
`dr_con("HH")` and `dr_con("HL")` and reflects the current state of the parquet files.

```{r}
#| label: surveys-covered
#| echo: true
#| eval: false

# Code used to generate the surveys table below.
library(obus)
library(dplyr)

hh <- dr_con("HH") |> collect()
hl <- dr_con("HL")

survey_stats <- hh |>
  filter(HaulValidity == "V", Survey != "NS-IBTS_UNIFtest") |>
  group_by(Survey) |>
  summarise(year_min = min(Year), year_max = max(Year), n_hauls = n(), .groups = "drop") |>
  arrange(desc(n_hauls))

species_per_survey <- hl |>
  filter(!is.na(Valid_Aphia), Valid_Aphia > 0) |>
  group_by(Survey) |>
  summarise(n_species = n_distinct(Valid_Aphia), .groups = "drop") |>
  collect()

survey_stats |> left_join(species_per_survey, by = "Survey")
```

| Survey | First year | Last year | Valid hauls | Species (AphiaIDs) |
|--------|-----------|----------|------------|-------------------|
| NS-IBTS | 1965 | 2026 | 36,566 | 837 |
| BTS | 1985 | 2025 | 19,887 | 1,012 |
| BITS | 1991 | 2026 | 18,659 | 193 |
| DYFS | 1985 | 2025 | 17,943 | 443 |
| Can-Mar | 1970 | 2021 | 11,372 | 408 |
| NSSS | 2008 | 2025 | 5,134 | 8 |
| SP-NORTH | 1990 | 2025 | 4,455 | 404 |
| EVHOE | 1997 | 2025 | 4,001 | 304 |
| IE-IGFS | 2003 | 2025 | 3,646 | 238 |
| FR-CGFS | 1988 | 2025 | 3,161 | 280 |
| SWC-IBTS | 1985 | 2010 | 2,702 | 174 |
| SNS | 1985 | 2025 | 2,138 | 230 |
| SP-PORC | 2001 | 2025 | 2,007 | 229 |
| NIGFS | 2005 | 2025 | 1,923 | 114 |
| SCOWCGFS | 2011 | 2026 | 1,790 | 168 |
| SP-ARSA | 1996 | 2025 | 1,732 | 246 |
| PT-IBTS | 2002 | 2023 | 1,551 | 392 |
| BTS-VIII | 2007 | 2025 | 1,437 | 113 |
| BTS-GSA17 | 2008 | 2025 | 1,128 | 8 |
| IE-IAMS | 2016 | 2025 | 1,105 | 192 |
| SCOROC | 2011 | 2025 | 636 | 90 |
| NL-BSAS | 2019 | 2025 | 408 | 2 |
| ROCKALL | 1999 | 2009 | 364 | 68 |
| FR-WCGFS | 2018 | 2025 | 361 | 97 |
| NS-IDPS | 2009 | 2016 | 250 | 21 |
| SE-SOUND | 2011 | 2025 | 155 | 50 |
| IS-IDPS | 2009 | 2009 | 89 | 67 |
| DWS | 2006 | 2009 | 71 | 173 |
| CODS-Q4 | 2024 | 2025 | 42 | 42 |

NS-IBTS is the largest survey (36,566 valid hauls from 1965); CODS-Q4 is the smallest (42
hauls from 2024). BTS records the most distinct species (1,012 AphiaIDs). `SWC-IBTS` ended
in 2010 and is superseded by `SCOWCGFS`.

#### What DATRAS metadata provides and does not provide

ICES provides field-level metadata through the DATRAS Technical Reference document, which
covers field names, types, units, and valid value ranges for all exchange table columns.
Survey-level metadata (gear specifications, time-series continuity, known breaks, QC history)
are maintained by the contributing national institutes and are not centrally available in
machine-readable form.

The IMBUS project addresses the gap in survey-level metadata under user requirement UR-Q1.
Concise per-survey fact sheets are not part of D2.2; they are treated as a separate deliverable.

#### Field and vocabulary reference {#sec-lookup}

{obus} exposes the DATRAS field-level and vocabulary metadata as two data objects:

**`dr_lookup_fields`** is the complete field dictionary for all DATRAS exchange and derived
product tables (HH, HL, CA, FL, LT, CPUEL, CPUEA, IDX). For each field it records the
standard name (`new`), the legacy name (`old`), the data type (`chr`/`int`/`dbl`), and a
plain-English description. The primary source is the ICES DATRAS web service endpoint
`getDatrasFieldList`
(`https://datras.ices.dk/WebServices/DATRASWebService.asmx/getDatrasFieldList`), which
covers the formal upload schema for HH, HL, CA, and FA. This is supplemented by
hand-curated entries for tables and fields not covered by the upload schema: LT and FL
(returned by `icesDatras` helper functions), the derived products CPUEL, CPUEA, and IDX,
and the `Valid_Aphia` / `aphia` mapping — a field added by the ICES datacenter during
processing that is absent from the formal upload field list. The script that builds
`dr_lookup_fields` is `data-raw/DATASET_lookup_fields.R` in the {obus} repository.

**`dr_lookup_vocabulary`** provides the controlled vocabularies for categorical DATRAS
fields — i.e. the valid codes and their descriptions for fields such as `HaulValidity`,
`DataType`, `LengthCode`, `Gear`, and `Country`. The source is the ICES Vocabulary service,
accessed via the `icesVocab` R package (`getCodeTypeList()` + `getCodeList()`). The full
ICES vocabulary is filtered to the subset of code types whose variable names correspond to
fields in `dr_lookup_fields`, giving one row per valid code value. The building script is
`data-raw/DATASET_vocabulary.R` in the {obus} repository.

```r
library(obus)

# Field dictionary: names, types, descriptions for all DATRAS tables
dr_lookup_fields

# Controlled vocabularies: valid codes for categorical fields
dr_lookup_vocabulary
```

These two objects together constitute the machine-readable metadata layer underlying the
field tables in §2. Both are regenerated by sourcing the respective `data-raw/` scripts
against live ICES web services; the versions shipped with {obus} reflect the state of
those services at the time of the last package build.

##### Controlled vocabulary tables {#sec-vocab}

The tables below list the valid codes for categorical fields. They are generated directly
from `dr_lookup_vocabulary` and are linked from the relevant rows in the field tables
above. Fields with large open-ended vocabularies (e.g. `Gear`, `Country`,
`StatisticalRectangle`) are excluded.

###### `HaulValidity` (HaulVal) {#vocab-haulval}

```{r}
dr_lookup_vocabulary |>
  filter(old == "HaulVal") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `DataType` {#vocab-datatype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "DataType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `ThermoCline` {#vocab-thermocline}

```{r}
dr_lookup_vocabulary |>
  filter(old == "ThermoCline") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `PelagicSamplingType` (PelSampType) {#vocab-pelsamtype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "PelSampType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `DoorType` {#vocab-doortype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "DoorType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `GearExceptions` (GearEx) {#vocab-gearex}

```{r}
dr_lookup_vocabulary |>
  filter(old == "GearEx") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `Rigging` {#vocab-rigging}

```{r}
dr_lookup_vocabulary |>
  filter(old == "Rigging") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

---

### Current Issues with DATRAS Data Formats and Meta-data Descriptions {#sec-issues}

#### Column naming convention

Two naming conventions exist in DATRAS, reflecting a change introduced by ICES around 2020:

| Convention | Source | Example fields |
|------------|--------|----------------|
| **Legacy** | `getDATRAS()` (XML API); pre-~2020 DATRAS downloads | `Ship`, `HaulNo`, `StNo`, `LngtCode`, `HLNoAtLngt`, `SubFactor`, `TotalNo`, `HaulVal`, `HaulDur` |
| **Standard** | Parquet files; IMBUS product; current ICES convention | `Platform`, `HaulNumber`, `StationName`, `LengthCode`, `NumberAtLength`, `SubsamplingFactor`, `TotalNumber`, `HaulValidity`, `HaulDuration` |

The co-existence of both conventions means that code written against one naming scheme fails
silently or with cryptic errors against the other. All field tables in §2.1 list both names.

#### The DataType field {#sec-datatype}

`DataType` is set at the haul level (HH) and governs how `NumberAtLength`, `TotalNumber`, and
`SpeciesCategoryWeight` in HL must be interpreted. It is the single most common source of
silent errors when working with exchange data, because the numeric meaning of HL fields depends
on a field in a different table.

| DataType | Sampling protocol | `SubsamplingFactor` | `NumberAtLength` | `n_haul` formula |
|----------|-------------------|---------------------|------------------|-----------------|
| `R` | Raw catch; some species sub-sampled | ≥ 1 | Count in sub-sample | `NumberAtLength × SubsamplingFactor` |
| `S` | Unsorted bulk sub-sample | > 1 | Count in sub-sample | `NumberAtLength × SubsamplingFactor` |
| `P` | Size-stratified (pseudocategory) | ≥ 1 per stratum | Count in sub-sample for stratum | `NumberAtLength × SubsamplingFactor` |
| `C` | Pre-raised to 1-hour haul | 1 | Numbers per hour (not a raw count) | `NumberAtLength × HaulDuration / 60` |

For DataTypes R, S, P: `n_hour = n_haul / HaulDuration × 60`  
For DataType C: `n_hour = NumberAtLength` (already per hour)

`TotalNumber` and `SpeciesCategoryWeight` are similarly DataType-dependent:

- **R/S/P:** total per-haul count and weight
- **C:** total per-hour count and weight (multiply by `HaulDuration / 60` to recover per-haul
  values)

**Size-stratified subsampling under DataType R.** The DATRAS schema intends DataType P for
hauls processed with size-stratified subsampling. In practice, GB-SCT vessels (293 groups
across NS-IBTS, SCOWCGFS, SCOROC) and DK vessels (82 groups from 2024 onward) submit
size-stratified hauls as DataType R. These are not data errors — the row-level formula
`NumberAtLength × SubsamplingFactor` remains correct. The risk arises only from direct use
of `TotalNumber`, which reflects the stratum total, not the species-haul total, in these
records.

```{r}
#| label: datatype-r-stratified
#| echo: true
#| eval: false

# Code used to derive the DataType R size-stratified statistics reported above.
library(obus)
library(dplyr)

hl  <- dr_con("HL")
hh  <- dr_con("HH")

grp <- c(".id", "Valid_Aphia", "IndividualSex", "DevelopmentStage", "SpeciesCategory")

# Groups where SubsamplingFactor varies within the natural group, under DataType R
sf_violations_R <- hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_sf = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_sf > 1) |>
  left_join(hh |> distinct(.id, DataType, Country), by = ".id") |>
  filter(DataType == "R")

sf_violations_R |> count()                          # 128 groups

sf_violations_R |> count(Country, sort = TRUE)      # GB-SCT: 97; DK: 31 (illustrative)
```

`DataType` must be joined from HH before any HL arithmetic:

```r
hl |> left_join(hh |> select(.id, DataType, HaulDuration), by = ".id")
```

#### Cross-table ambiguities

Several fields share names or concepts across tables but differ in meaning.

**Across HL and CA:**

| Issue | HL | CA |
|-------|----|----|
| Sex field | `SpeciesSex` (`IndividualSex` in parquet) — visual, population-level | `IndividualSex` — determined by dissection |
| `NumberAtLength` | Sub-sample count per length bin; multiply by `SubsamplingFactor` for population estimate | Count of aged fish at that length class; used as raising key for ALK |
| `LengthClass` | Lower boundary of a length frequency bin | Measured length of an individual fish |

**Gear fields duplicated across HH, HL, and CA:**

`Gear`, `GearExceptions`, `SweepLength`, and `DoorType` appear in all three tables with
identical values. This is a submission artefact. HH is the sole source of truth; the HL and
CA copies are redundant after joining on `.id`.

#### Other known data issues

- **`SubsamplingFactor = NA`:** Approximately 0.015% of HL records have missing
  `SubsamplingFactor` under DataType R (Norwegian and Dutch vessels in NS-IBTS). This
  propagates as `NA` in computed `n_haul` and `n_hour`.
- **`TotalNumber` rounding for DataType C:** `TotalNumber` is a back-converted CPUE count;
  arithmetic checks reveal rounding differences of 1–20 fish (~1% of DataType C groups). This
  is an artefact of the ICES submission process and is expected.
- **CA `AreaCode`/`AreaType`:** CA rows can represent area-pooled age samples, not individual
  haul observations. Check `AreaType` before joining CA to HH.
- **Area standardisation:** Mapping from haul positions to C-squares and ICES divisions is
  partially implemented (`dr_assign_area()`) but lookup helpers are not yet complete in {obus}.

---

::: {.callout-note appearance="minimal"}
**End of D2.2 §2 — DATRAS current format documentation.**  
The sections above (§1–2) constitute the DATRAS data format and meta-data reference. The
following section (§3) specifies the IMBUS standardised product and serves as the technical
specification for the D2.3 and D2.4 data deliverables.
:::

## IMBUS Proposed Data Formats — D2.3 Product Specification {#sec-imbus}

> **Deliverable D2.3 scope.** This section is the technical specification for the IMBUS
> integrated data product (D2.3, Type: DATA, due M12). It defines the proposed column
> names, file formats, data tier structure, and code pathways. The D2.3 product itself —
> the parquet files — will be hosted on the ICES open access portal under D2.4 (M12).
> The current interim location is <https://heima.hafro.is/~einarhj/datras/>.
>
> Long-term sustainability is handled by WP5: **D5.3** (ICES, M18) formally integrates
> the IMBUS data products and the {obus} codebase into the ICES DATRAS R library and
> ICES web services, ensuring D2.3 has an institutional home beyond the project period.
> The raw DATRAS data supply that feeds D2.3 is formalised in **D5.1** (ICES, M6).

This section describes the IMBUS proposals for standardising the DATRAS data formats, the
file formats to be used for distribution, how meta-data descriptions will be handled, and
the standardised code provided for processing DATRAS data.

### Standardized Data Formats {#sec-std-formats}

#### Harmonized variables

The IMBUS product uses **standard column names throughout** (see [Column naming convention]).
Where data originate from the legacy XML API, an automatic translation step maps legacy to
standard names using the field dictionary `dr_lookup_fields`. Fields with no standard
equivalent retain their original names. The full mapping (~292 fields) is available in the
`dr_lookup_fields` data object in {obus} and in the companion
[DATRAS Field Reference](datras_field_reference.qmd).

Additional standardisation steps applied to all exchange tables:

- **Sentinel value removal.** The DATRAS convention of using −9 for missing values is replaced
  with proper `NA`.
- **Species resolution.** Historical species codes (ITIS TSN, NODC) are mapped to current
  WoRMS AphiaIDs (`ValidAphiaID`) for cross-era joins.
- **Unique haul identifier.** An 8-field composite key (`.id`) uniquely identifies each haul
  across all surveys, years, and countries.

#### Unique Haul Identifier {#unique-haul-identifier}

`dr_add_id()` constructs a composite key that uniquely identifies each haul:

```
.id = Survey:Year:Quarter:Country:Platform:Gear:StationName:HaulNumber
```

Example: `"NS-IBTS:2023:1:NL:58G2:GOV:18:27"`

This is the recommended join key for all operations. A shorter 6-field key
(`.id2 = Survey:Year:Quarter:Platform:Gear:HaulNumber`) is also available for comparison
with the ICES CPUEL derived product, which omits `Country` and `StationName`.

`dr_add_id()` auto-detects legacy vs. standard column names.

#### Proposed HL split design {#hl-split}

The preferred end-state is to split `HL.parquet` into two separate files, separating catch
totals from length frequencies. This requires confirmation from Vaishav Soni (ICES) and the
IMBUS team on the field gaps noted in [Sampling protocol] (action point A2, 1 June 2026).

Open questions:

- Should `DevelopmentStage` be part of the grouping key, or aggregated away?
- How to handle DataType P hauls where the same species appears in multiple `SpeciesCategory` strata?
- Should `SpeciesSex` be retained in the split tables or aggregated?
- What is the `SpeciesCategoryWeight` definition when `DevelopmentStage` varies within a group?

Until the split is agreed, `HL.parquet` is distributed as a single flat file. The
`dr_cpue_by_haul(diag = TRUE)` function approximates the intended catch-level table;
`dr_cpue_by_length(diag = TRUE)` approximates the length-frequency table.

**`HL_catch.parquet`** (proposed) — one row per `.id` × `ValidAphiaID` × `SpeciesSex` ×
`DevelopmentStage` × `SpeciesCategory`:

| Field | Type | Unit | Description |
|-------|------|------|-------------|
| `.id` | chr | — | Haul join key |
| `ValidAphiaID` | int | — | Resolved WoRMS AphiaID |
| `SpeciesValidity` | chr | — | Record type filter (see [Sampling protocol](#hl-sampling-protocol)) |
| `SpeciesSex` | chr | — | Population-level sex (visual) |
| `DevelopmentStage` | chr | — | Non-invasive stage code |
| `SpeciesCategory` | int | — | Subsampling category ID |
| `SubsamplingFactor` | dbl | — | Raise factor (retained for QC) |
| `SubsampledNumber` | int | count | Specimens actually measured |
| `SubsampleWeight` | int | g | Weight of measured subsample |
| `n_haul` | dbl | count | DataType-corrected total catch per haul (replaces `TotalNumber`) |
| `w_haul` | dbl | g | DataType-corrected catch weight per haul (replaces `SpeciesCategoryWeight`) |

**`HL_length.parquet`** (proposed) — one row per `.id` × `ValidAphiaID` × `SpeciesSex` ×
`DevelopmentStage` × `SpeciesCategory` × `length_cm`:

| Field | Type | Unit | Description |
|-------|------|------|-------------|
| `.id` | chr | — | Haul join key |
| `ValidAphiaID` | int | — | Resolved WoRMS AphiaID |
| `SpeciesSex` | chr | — | Population-level sex |
| `DevelopmentStage` | chr | — | Non-invasive stage |
| `SpeciesCategory` | int | — | Subsampling category |
| `length_cm` | dbl | cm | Standardised length (from `LengthCode` + `LengthClass`) |
| `LengthType` | chr | — | Measurement type (total, fork, mantle, etc.) |
| `NumberAtLength` | dbl | count | Sub-sample count at this length class |
| `SubsamplingFactor` | dbl | — | Retained for scaling to haul-level counts |

**`CA_raised.parquet`** (exploratory — requires further design) — one row per `.id` × `aphia`
× `age`:

| Field | Type | Unit | Description |
|-------|------|------|-------------|
| `.id` | chr | — | Haul join key |
| `aphia` | int | — | Resolved WoRMS AphiaID |
| `age` | int | years | Age class |
| `plus_group` | lgl | — | Whether this row represents the open-ended plus group |
| `n_haul` | dbl | count | Estimated numbers at age per haul (from ALK × HL catch) |
| `n_hour` | dbl | count/hr | Numbers at age per hour of hauling |
| `alk_source` | chr | — | Identifies the pooling level used to construct the ALK |

Design of `CA_raised` requires input from WP4 and resolution of ALK pooling level, plus-group
threshold, CA coverage gaps, and area-pooled CA rows before implementation can proceed.

#### Summary: proposed data tiers

```
Tier 0 — Raw exchange parquets
  HH.parquet, HL.parquet, CA.parquet, LT.parquet
  All original submission fields; legacy names translated to standard on fetch.
  For: expert users, QC, full-fidelity reproduction.

Tier 1 — Intermediate (middle-data) parquets      ← proposed
  HL_catch.parquet    DataType-corrected counts and weights by haul/species/category
  HL_length.parquet   Standardised length (cm) and sub-sample counts
  CA_raised.parquet   Age-disaggregated abundance (ALK-based; design TBD)
  For: most analytical users; no knowledge of DataType or LengthCode required.

Tier 2 — Derived products
  CPUE-by-length  dr_cpue_by_length() applied to Tier 1 or Tier 0
  CPUE-by-age     User ALK pipeline applied to Tier 1 (exploratory)
  CPUEA / IDX     Official ICES products; accessible via dr_get()
  For: downstream modelling, stock assessment, spatial analysis.
```

---

### File Formats Used {#sec-file-formats}

#### Parquet

The IMBUS product distributes HH, HL, and CA as columnar **parquet** files. Parquet enables
fast, column-selective querying without downloading the full table — a critical advantage
given that HL alone contains tens of millions of rows across all surveys and years.

| Table | URL |
|-------|-----|
| HH | `https://heima.hafro.is/~einarhj/datras/HH.parquet` |
| HL | `https://heima.hafro.is/~einarhj/datras/HL.parquet` |
| CA | `https://heima.hafro.is/~einarhj/datras/CA.parquet` |

These files are readable by any tool that supports parquet or DuckDB: R, Python
(`duckdb`, `pyarrow`, `pandas`), Julia, DBeaver, and others.

#### Data access via {obus}

**Lazy parquet connection (recommended):**

```r
library(obus)
library(dplyr)

# Connect without downloading — filter then collect
hh <- dr_con("HH") |> filter(Survey == "NS-IBTS", Year >= 2020)
hl <- dr_con("HL") |> filter(Survey == "NS-IBTS", Year >= 2020)
ca <- dr_con("CA") |> filter(Survey == "NS-IBTS", Year >= 2020)

hh_local <- collect(hh)
```

`dr_con()` returns a lazy DuckDB connection. Filters are pushed down to the parquet file;
only the selected rows and columns are transferred into R memory.

**Full download via `dr_get()`:**

```r
# All surveys, all years — downloads entire table
hh <- dr_get("HH")

# Filtered via legacy XML API (slower; use parquet for large queries)
hh <- dr_get("HH", surveys = "NS-IBTS", years = 2020:2024, source = "xml")
```

#### {obus}: interim status and transition path {#sec-obus-status}

{obus} is **explicitly experimental** (lifecycle status: experimental) and is best understood
as the proof-of-concept implementation for the WP2 data standardisation work. It is not
intended to be a permanent, standalone R package. Its role within the IMBUS project lifecycle
is:

1. **Proof-of-concept.** Provide a working, validated implementation of the naming
   conventions, DataType arithmetic, parquet access, and CPUE pathways described in this
   document during the WP2 delivery period (M1–M12).
2. **D2.3 delivery vehicle.** Serve as the primary means of accessing the D2.3 data product
   (parquet files) while those files are hosted on the interim server
   (<https://heima.hafro.is/~einarhj/datras/>). The server is experimental; simultaneous
   traffic capacity is untested.
3. **Design feedback loop.** Accumulate community experience and feedback that directly
   informs what is worth upstreaming into the official ICES infrastructure.

{obus} **builds on, rather than replaces**, the official `icesDatras` package — it imports
`icesDatras` and delegates XML-source downloads to `icesDatras::getDATRAS()`. The added
value is the parquet/DuckDB access layer, unified column naming (`dr_translate()`),
DataType-aware derived fields (`dr_add_n_and_cpue()`, `dr_add_length_mm()`), and the
quality-check family (`dr_check_*()`).

**Intended trajectory.** Under WP5/D5.3 (ICES, M18), the functionality proven in {obus} is
a candidate for integration into the official ICES DATRAS R library (`icesDatras`), alongside
migration of the parquet files to ICES-managed infrastructure. At that point {obus} will
either be archived or retained as a lightweight convenience wrapper around the official
package. The naming-convention mapping (`dr_lookup_fields`), the DataType arithmetic, and the
composite haul key (`.id`) are the most transferable elements; the DuckDB/parquet access
layer will follow once the files move to ICES hosting (D2.4 / D5.3).

---

### Updated Meta-data Descriptions {#sec-updated-metadata}

The IMBUS standard names and field definitions described in §2 and §3.1 constitute an updated
meta-data description for the DATRAS exchange tables. The companion document
[DATRAS Field Reference](datras_field_reference.qmd) provides complete field reference tables
for all HH, HL, CA, and LT fields, including types, units, valid values, and a drop-or-derive
quick reference.

Per-survey meta-data fact sheets (coverage, gear history, time-series breaks, QC history) are
not produced by D2.2. These are addressed separately under user requirement UR-Q1 and are
expected to be a deliverable in a later phase of WP2.

---

### Standardized Code for Processing DATRAS Data {#sec-code}

WP2 does not generate new CPUE data series. Instead, it provides transparent, reproducible
code that any user can apply to the standardised exchange tables to generate CPUE on the fly.

#### Data flow pipeline

```
DATRAS database
      │
      ▼
Raw exchange download    ← getDATRAS() (legacy XML API)
      │
      ▼
Type normalisation       ← dr_settypes(): column types enforced; −9 → NA
Column renaming          ← dr_translate(): legacy → standard names
      │
      ▼
Parquet files (server)   ← HH.parquet / HL.parquet / CA.parquet
      │
      ▼
User pipeline            ← dr_con() / dr_get() → filter → collect()
                         → dr_add_id(), dr_add_length_mm(), dr_add_n_and_cpue(), …
```

#### {obus} helper functions

The R package {obus} provides helper functions that add derived columns to HL after
collection:

| Function | Output column(s) | Description |
|----------|-----------------|-------------|
| `dr_add_id()` | `.id` | 8-field composite haul key |
| `dr_add_length_cm()` | `length_cm` | Length in cm (standardised from `LengthCode` + `LengthClass`) |
| `dr_add_length_mm()` | `length_mm` | Length in mm |
| `dr_add_n_and_cpue()` | `n_haul`, `n_hour` | DataType-corrected counts per haul and per hour |
| `dr_add_record_type()` | `record_type` | Integer code classifying the HL record pattern |

`dr_add_n_and_cpue()` computes `n_haul` as `NumberAtLength × SubsamplingFactor` at the
individual row level for DataTypes R, S, and P, and as `NumberAtLength × HaulDuration / 60`
for DataType C. It never reads `TotalNumber` directly and is therefore safe for
size-stratified hauls regardless of DataType coding.

#### Zero observations

A CPUE value of zero means the species was **absent** from that haul but was caught by at
least one vessel in the same survey/year/quarter. Zero rows are distinct from `NA` (missing
data) and are required for unbiased abundance indices.

`dr_cpue_by_length(zerofill = TRUE)` and `dr_cpue_by_haul(zerofill = TRUE)` insert explicit
zero rows for all haul × species combinations where the species is absent from the haul but
present elsewhere in the survey/year/quarter. Zero-fill adds substantial rows and forces
collection into memory; it is not applied by default.

#### CPUE calculation pathways {#cpue-pathways}

**CPUE-by-length** requires five steps:

1. **Filter valid hauls:** retain `HaulValidity == "V"`.
2. **Filter standard species records:** retain `SpeciesValidity == "1"`.
3. **Join DataType and HaulDuration from HH to HL** on `.id`.
4. **Compute `n_haul` and `n_hour`** using the DataType-appropriate formula.
5. **Aggregate** across `SpeciesSex` and `SpeciesCategory` (sum `n_hour` per `.id` ×
   `ValidAphiaID` × `length_mm`).

```r
library(obus)
library(dplyr)

hh <- dr_con("HH") |> filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1)
hl <- dr_con("HL") |> filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1)

cpue    <- dr_cpue_by_length(hh, hl)
cpue_zf <- dr_cpue_by_length(hh, hl, zerofill = TRUE)
```

Output columns: `.id`, `Survey`, `Year`, `Quarter`, `ValidAphiaID`, `length_mm`, `n_hour`.
Zero-fill rows have `length_mm = NA` and `n_hour = 0`.

The output has been validated against the official ICES CPUEL product
(`dr_get("CPUEL", ...)`).

**Exploratory CPUE-by-age** is achievable from the exchange tables but involves analytical
choices not standardised here:

| | Official CPUEA | Exploratory CPUE-by-age |
|-|---------------|------------------------|
| ALK source | ICES expert-group agreed | User-defined from CA |
| Pooling level | Survey/species specific | User's choice |
| Plus-group handling | Agreed convention | User's choice |
| Appropriate for | Formal stock assessment | Research, exploration |
| Accessible via | `dr_get("CPUEA", ...)` | User pipeline on HL + CA |

The general pathway: (1) compute CPUE-by-length; (2) construct ALK from CA; (3) apply ALK to
length CPUE; (4) aggregate across hauls. The proposed Tier 1 parquets ([Proposed HL split
design]) would make step 1 trivial, reducing the entry barrier for this approach.

---

## Discussion and Conclusions {#sec-discussion}

*[To be completed.]*

---

## Further Work {#sec-further-work}

The following items are open as of the date of this document:

1. **HL split design.** The split of `HL.parquet` into `HL_catch` and `HL_length` parquets
   cannot be finalised until field gaps (`SubsampledNumber`, `SubsampleWeight`,
   `DevelopmentStage`, `LengthType` semantics) are confirmed by Vaishav Soni and Adrianna
   Villamor (action point A2, 1 June 2026).

2. **CA section.** Field-by-field treatment of CA comparable to the HH and HL sections is
   needed; the current CA section is a stub.

3. **Parquet hosting.** The current parquet files are on a personal server
   (`heima.hafro.is`). Transfer to the ICES open access portal is pending WP5/ICES
   coordination (D2.4).

4. **Integration into ICES infrastructure (D5.3).** Per decisions D2 and D3 from the
   1 June 2026 IMBUS meeting, parquet/DuckDB functionality should be incorporated into
   the official ICES DATRAS R package rather than maintained as a separate tool. This is
   formally the scope of WP5 deliverable D5.3 (ICES, M18), which integrates the D2.3
   data product and {obus} codebase into ICES web services and the DATRAS R library.

5. **Per-survey meta-data fact sheets.** Coverage, gear history, time-series breaks, and QC
   caveats at the survey level (user requirement UR-Q1) are a separate deliverable not
   covered by D2.2.

6. **Area standardisation.** Mapping from haul positions to C-squares and ICES divisions is
   partially implemented (`dr_assign_area()`) but the lookup helpers are not yet complete.

---

## References {#sec-references}

*[To be completed.]*

- ICES DATRAS Technical Reference (available from the ICES Datacenter)
- IMBUS D2.1 — User requirements (April 2026)
- WoRMS Editorial Board (2024). World Register of Marine Species. <https://www.marinespecies.org>

---

## Annex 1 — Related Documents and Resources {#sec-annex}

The following companion resources are part of the IMBUS WP2 output but are not reproduced
in the body of this deliverable:

- **[DATRAS Field Reference](datras_field_reference.qmd)** — Complete field reference tables
  for all HH, HL, CA, and LT columns; length-unit and species-coding rules; area
  standardisation; data quality checks; drop-or-derive quick reference; {obus} code quick
  reference.
- **`docs_external/DATRAS_Technical_Reference.md`** — Full DATRAS field and unit reference
  (ICES).
- **`background/IMBUS_proposal_summary.qmd`** — IMBUS project context and WP2 objectives.
- **`meetings/20260601_WP2-WP5.qmd`** — Minutes of the WP2-WP5 alignment meeting,
  1 June 2026.
