---
title: "D2.2 — Data Formats and Metadata"
subtitle: "IMBUS Deliverable D2.2 · Work Package 2"
author: "Einar Hjörleifsson (subcontracted to DTU Aqua)"
date: "July 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    embed-resources: true
  docx:
    reference-doc: ../reference.docx
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(obus)
library(dplyr)
library(knitr)
```

> **Project:** IMBUS — Implementing More and Better Use of ICES Survey Data  
> **Deliverable:** D2.2 — Data formats and meta-data  
> **Work Package:** WP2 — User requirements, data formats and meta-data  
> **Status:** DRAFT — circulated for review  
> **Due:** Month 11 (July 2026)

---

## Executive Summary

This document is the formal output of Task T2.3 (Data Formats and Meta-data) of the IMBUS
project. It documents the current ICES DATRAS exchange format and its metadata — providing
a field-by-field reference for the **HH** (haul header), **HL** (haul-level catch at
length), and **CA** (catch at age) record types.

The scope of D2.2 is the exchange data itself: the current field names, types, units,
valid values, known issues, and the survey coverage available through DATRAS. The proposed
IMBUS standardised data product (parquet files, derived CPUE pathways, intermediate data
tiers) is specified in the companion deliverable **D2.3**
([D2.3 — IMBUS Integrated Data Product Specification](D2.3_imbus_product_specification.qmd),
due August 2026).

The principal documentation findings are:

1. **Two co-existing naming conventions.** DATRAS field names changed around 2020
   (e.g. `Ship` → `Platform`, `HaulNo` → `HaulNumber`). Both conventions are documented
   throughout, with legacy names shown in parentheses.
2. **HL structure.** The HL table combines haul join keys, species identification, sampling
   protocol bookkeeping, catch totals, and length frequencies in a single flat structure.
   These are documented as distinct field groups with known interpretation pitfalls.
3. **Sentinel value −9.** Used throughout DATRAS for missing or inapplicable fields; not
   a real value.
4. **Species resolution via WoRMS AphiaID.** Historical species codes (ITIS TSN, NODC) are
   mapped to current WoRMS AphiaIDs in the field `ValidAphiaID` (`Valid_Aphia`).
5. **DataType field.** The `DataType` field in HH governs how several HL fields must be
   interpreted; it is the most common source of silent errors in user code.

---

## Background and Purpose {#sec-background}

The ICES DATRAS database holds fisheries-independent bottom trawl survey data submitted by
national institutes across the Northeast Atlantic, covering surveys from 1965 to the present.
The data are structured as three exchange tables — HH, HL, and CA — which together describe
each trawl haul, its catch by length class, and biological measurements on individual fish.

WP2 has four objectives:

1. Assess the user requirements of different user groups with respect to potential applications
   of survey data. *(Addressed in D2.1, April 2026.)*
2. Develop a data format that can be used to generate the products described in WP3–5.
3. Generate meta-data descriptions of the overall data.
4. Process the current data available in the DATRAS database into the new format, while
   correcting for any errors detected and taking account of zero observations and
   length/weight conversions.

Objective 1 has been delivered in D2.1 (IMBUS user requirements, April 2026, n = 55
respondents), which identified four dominant user challenges:

* Complex table structure (HH/HL/CA hierarchy and join logic)
* Column name changes introduced around 2020 (legacy names vs. "new" names)
* Quality-control inconsistencies
* Slow downloads

The current deliverable D2.2 addresses objectives 2 and 3 — the current data formats and
the associated meta-data descriptions. D2.2 covers the three exchange tables (HH, HL, CA),
their field-level metadata, and known issues with the current format. The proposed IMBUS
standardised data product (parquet files, CPUE pathways, unique haul identifier) is covered
in D2.3. A full scope statement is in [Appendix 2](#sec-appendix2).

---

## ICES DATRAS Data Formats and Meta-data Descriptions {#sec-datras}

The ICES DATRAS database distributes bottom trawl survey data as three exchange tables. Each
table is submitted independently per survey, year, quarter, and country; the tables are
linked by a common set of haul identification fields. The following sections describe the
current format of each table field by field, the metadata available for the surveys covered,
and known issues with the current formats.

The field specifications below are drawn from two machine-readable metadata sources
maintained in {obus}: `dr_lookup_fields` (field names, types, and descriptions — see
[Field and vocabulary reference]) and `dr_lookup_vocabulary` (controlled vocabularies for
categorical fields — see [Field and vocabulary reference]).

### Data Formats {#sec-formats}

#### HH — Haul Header {#sec-hh}

One row per haul. HH is the parent table; HL and CA join to it on the eight common haul
identification fields listed below.

##### Haul identification and join key

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `Survey` | `Survey` | chr | Survey acronym (e.g. `"NS-IBTS"`) |
| `Year` | `Year` | int | Year of survey |
| `Quarter` | `Quarter` | int | Quarter (1–4) |
| `Country` | `Country` | chr | ISO country code of contributing vessel |
| `Platform` | `Ship` | chr | Vessel identifier (SeaDataNet code) |
| `Gear` | `Gear` | chr | Gear code (e.g. `"GOV"`) |
| `StationName` | `StNo` | chr | Station identifier within survey |
| `HaulNumber` | `HaulNo` | int | Haul sequence number |
| `.id` | — | chr | Composite haul key added by {obus}: `Survey:Year:Quarter:Country:Platform:Gear:StationName:HaulNumber` — see [D2.3](D2.3_imbus_product_specification.qmd#unique-haul-identifier) for full specification |

##### Haul outcome

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `HaulValidity` | `HaulVal` | chr | — | `"V"` = valid; `"I"` = invalid. Filter to `"V"` for standard analyses — [all codes](#vocab-haulval) |
| `DataType` | `DataType` | chr | — | Sampling protocol (see [Current Issues](#sec-issues)) — [codes](#vocab-datatype) |
| `HaulDuration` | `HaulDur` | int | Minutes | Duration of tow |
| `ReasonHaulDisruption`^[Specified in ICES vocabulary but not in the upload schema; no legacy equivalent] | — | chr | — | Disruption code for invalid hauls |

##### Temporal

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `Month` | `Month` | int | — | Month of haul |
| `Day` | `Day` | int | — | Day of haul |
| `StartTime` | `TimeShot` | chr | HHMM GMT | Start time of haul in HHMM format (GMT); combine with `Year`, `Month`, `Day` for a full datetime |

##### Position and area

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `ShootLatitude` | `ShootLat` | dbl | Decimal degrees | Tow start latitude |
| `ShootLongitude` | `ShootLong` | dbl | Decimal degrees | Tow start longitude |
| `HaulLatitude` | `HaulLat` | dbl | Decimal degrees | Tow end latitude |
| `HaulLongitude` | `HaulLong` | dbl | Decimal degrees | Tow end longitude |
| `StatisticalRectangle` | `StatRec` | chr | — | ICES statistical rectangle; derivable from (start) position |
| `SurveyIndexArea`^[Specified in ICES vocabulary but not in table] | `SurveyIndexArea` | chr | — | Survey-specific stratum used in index calculations |
| `DepthStratum` | `DepthStratum` | chr | — | Survey-specific depth band |
| `BottomDepth` | `Depth` | int | m | Seabed depth at the haul location, measured from water surface |
| `DayNight` | `DayNight` | chr | — | `"D"` / `"N"` (15 min before/after sunrise/sunset) |

`StatisticalRectangle` and `DayNight` are derivable from position and time; they are retained
for convenience. `BottomDepth` is a property of the station location; it also determines
`WarpLength` in the gear setup.

##### Gear fields {#hh-gear}

Fields describe the tow and gear configuration. `SweepLength`, `DoorType`, and
`GearExceptions` are also present in HL and CA as a submission artefact; HH is the
authoritative source.

**Tow parameters** (how the haul was executed)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `Distance` | `Distance` | dbl | m | Distance between haul start and end |
| `SpeedGround` | `GroundSpeed` | dbl | knots | Ground speed of towing |
| `SpeedWater` | `SpeedWater` | dbl | knots | Trawl speed through water |
| `TowDirection` | `TowDir` | int | degrees | Towing direction; 360 = N |
| `MinTrawlDepth` | `MinTrawlDepth` | int | m | Shallowest point of trawl path (pelagic hauls) |
| `MaxTrawlDepth` | `MaxTrawlDepth` | int | m | Deepest point of trawl path (pelagic hauls) |

**Gear geometry** (sensor measurements taken during the haul)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `WingSpread` | `WingSpread` | dbl | m | Mean wing spread |
| `DoorSpread` | `DoorSpread` | dbl | m | Mean door spread |
| `NetOpening` | `Netopening` | dbl | m | Mean vertical net opening |

**Gear configuration** (rigging set before deployment)

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SweepLength` | `SweepLngt` | int | m | Sweep length; survey- and quarter-specific recommended values |
| `DoorType` | `DoorType` | chr | — | Type of door used — [codes](#vocab-doortype) |
| `DoorSurface` | `DoorSurface` | dbl | m² | Door surface area |
| `DoorWeight` | `DoorWgt` | int | kg | Door weight |
| `WarpLength` | `Warplngt` | int | m | Warp length; determined by `BottomDepth` |
| `WarpDiameter` | `Warpdia` | int | mm | Warp diameter |
| `WarpDensity` | `WarpDen` | int | kg/m | Warp weight per linear metre |
| `Rigging` | `Rigging` | chr | — | Rigging method; survey-specific — [codes](#vocab-rigging) |
| `Tickler` | `Tickler` | int | — | Number of tickler chains |
| `Buoyancy` | `Buoyancy` | int | kg | Total buoyancy of net floats |
| `KiteArea` | `KiteDim` | dbl | m² | Kite area (pelagic hauls) |
| `GroundRopeWeight` | `WgtGroundRope` | int | kg | Total groundrope weight |
| `CodendMesh` | `CodendMesh` | int | mm | Codend mesh size |
| `PelagicSamplingType` | `PelSampType` | chr | — | Pelagic trawl sampling design — [codes](#vocab-pelsamtype) |
| `GearExceptions` | `GearEx` | chr | — | Deviations from standard gear; survey-specific — [codes](#vocab-gearex) |

##### Environmental fields {#hh-env}

Fields describe oceanographic and meteorological conditions at the time of the haul.
Direction fields use the convention 0 = no movement, 360 = flow/movement from south to
north, −1 = varying direction.

**Water column**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SurfaceTemperature` | `SurTemp` | dbl | °C | Surface water temperature |
| `BottomTemperature` | `BotTemp` | dbl | °C | Bottom water temperature |
| `SurfaceSalinity` | `SurSal` | dbl | PSU | Surface salinity |
| `BottomSalinity` | `BotSal` | dbl | PSU | Bottom salinity |
| `ThermoCline` | `ThermoCline` | chr | — | Thermocline observed? (`"Y"` / `"N"`); defined as ≥ 2 °C change over 10 m — [codes](#vocab-thermocline) |
| `ThermoClineDepth` | `ThClineDepth` | int | m | Depth of thermocline; populated only when `ThermoCline == "Y"` |
| `SecchiDepth` | `SecchiDepth` | dbl | m | Secchi disc depth (water clarity) |
| `Turbidity` | `Turbidity` | int | NTU | Turbidity |

**Currents and tides**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `SurfaceCurrentDirection` | `SurCurDir` | int | degrees | Surface current direction |
| `SurfaceCurrentSpeed` | `SurCurSpeed` | dbl | m/s | Surface current speed |
| `BottomCurrentDirection` | `BotCurDir` | int | degrees | Bottom current direction |
| `BottomCurrentSpeed` | `BotCurSpeed` | dbl | m/s | Bottom current speed |
| `TidePhase` | `TidePhase` | int | min | Minutes before next high tide |
| `TideSpeed` | `TideSpeed` | dbl | m/s | Tidal speed |

**Wind and swell**

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `WindDirection` | `WindDir` | int | degrees | Wind direction; calm = 0, varying = −1 |
| `WindSpeed` | `WindSpeed` | int | m/s | Wind speed |
| `SwellDirection` | `SwellDir` | int | degrees | Swell direction; varying = −1 |
| `SwellHeight` | `SwellHeight` | dbl | m | Swell/wave height |

---

#### HL — Haul-level Length {#sec-hl}

HL contains one row per length class per species per haul. It is the most complex of the three
exchange tables, combining haul join keys, species identification, sampling protocol
bookkeeping, catch totals, and length-frequency distributions in a single flat structure.
Understanding how these groups of fields interact is essential for correct CPUE arithmetic.

##### Haul identification and join key {#hl-haul-id}

HL shares the same eight haul-identification fields as HH. After joining HL to HH on `.id`,
these fields are redundant in HL and can be dropped.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `Survey` | `Survey` | chr | Survey acronym (e.g. `"NS-IBTS"`) |
| `Year` | `Year` | int | Year of survey |
| `Quarter` | `Quarter` | int | Quarter (1–4) |
| `Country` | `Country` | chr | ISO country code of contributing vessel |
| `Platform` | `Ship` | chr | Vessel identifier (SeaDataNet code) |
| `Gear` | `Gear` | chr | Gear code (e.g. `"GOV"`) |
| `StationName` | `StNo` | chr | Station identifier within survey |
| `HaulNumber` | `HaulNo` | int | Haul sequence number |
| `.id` | — | chr | Composite haul key added by {obus} — see [D2.3](D2.3_imbus_product_specification.qmd#unique-haul-identifier) |
| `RecordHeader` | `RecordType` | chr | Record type identifier; always `"HL"` |

Three gear fields are duplicated from HH as a submission artefact (each exchange file is
self-describing). HH is the sole source of truth; the HL copies are redundant after joining
on `.id`.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `GearExceptions` | `GearEx` | chr | Non-standard gear modifications — [codes](#vocab-gearex) |
| `SweepLength` | `SweepLngt` | int | Sweep length (metres) |
| `DoorType` | `DoorType` | chr | Door type code — [codes](#vocab-doortype) |

##### Species identification {#hl-species-id}

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `ValidAphiaID` | `Valid_Aphia` | int | Resolved WoRMS AphiaID; primary species join key across HL, CA, and external taxonomy tables |
| `SpeciesCode` | `SpecCode` | chr | Original species code as submitted (ITIS TSN, NODC, or WoRMS) |
| `SpeciesCodeType` | `SpecCodeType` | chr | Code system used: `"TSN"`, `"NODC"`, `"WoRMS"` — [codes](#vocab-speccodeype) |
| `DateofCalculation` | `DateofCalculation` | chr | Date on which the ICES datacenter resolved `SpeciesCode` to `ValidAphiaID` |

`SpeciesCode` and `SpeciesCodeType` reflect the original submission; `ValidAphiaID` is the
resolved, standardised identifier and should be used for all joins and aggregations.

##### Sampling protocol {#hl-sampling-protocol}

These fields describe how the catch was processed for a given species × sex × category group.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `SpeciesValidity` | `SpecVal` | chr | Record type — governs which rows to include in products (see below) — [codes](#vocab-specval) |
| `SpeciesCategory` | `CatIdentifier` | chr | Subsampling category ID; up to 5 per species/sex/haul; always `"1"` for DataType C |
| `SubsamplingFactor` | `SubFactor` | dbl | Raise factor: e.g. 6 means 1 in 6 fish was measured |
| `SubsampledNumber` | `NoMeas` | dbl | Count of fish actually measured in the subsample |
| `SubsampleWeight` | `SubWgt` | dbl | Weight of the measured subsample (grams) |

> **Note (A2):** `SubsampledNumber` and `SubsampleWeight` are fields not confirmed in all
> surveys. Vaishav Soni and Adrianna Villamor have been asked to confirm their presence and
> exact semantics in the exchange data (action point A2, 1 June 2026).

**`SpeciesValidity`** controls what data are present in a row and which rows should be
included in derived products:

| Value | Meaning | Include in products |
|-------|---------|---------------------|
| `"1"` | Full record: lengths, numbers, weights | Yes |
| `"0"` | Invalid | No |
| `"4"` | Total number only; no lengths | Situation-dependent |
| `"5"` | Presence/absence only | Situation-dependent |
| `"6"` | Weight only | Situation-dependent |
| `"7"` | Numbers and weights; no lengths | Situation-dependent |
| `"10"` | Lengths and numbers; no weights | Situation-dependent |

Standard DATRAS products filter to `SpeciesValidity == "1"`.

**Constancy of sampling protocol fields.** These fields are in principle constant within a
natural group (`.id` × `ValidAphiaID` × `SpeciesSex` × `DevelopmentStage` × `SpeciesCategory`),
but empirical checking reveals two important exceptions:

- **`SpeciesValidity`** varies within 2,786 of the ~2.6 million total groups (0.1%),
  predominantly BTS (1,289 groups) and Can-Mar (1,272 groups). The most common
  co-occurrences are `{1, 5}` (1,281 groups) and `{4, 7}` (580 groups), reflecting the
  coexistence of full length-frequency rows and supplementary rows (totals-only,
  presence/absence) within the same group. This is by design in those surveys.

- **`SubsamplingFactor`** varies within 389 of the ~2.1 million valid groups (0.02%) due to
  **size-stratified subsampling** — small and large fish are subsampled at different rates
  within the same species group. This pattern occurs predominantly in GB-SCT vessels (293
  groups, across NS-IBTS, SCOWCGFS, SCOROC) and DK vessels (82 groups, from 2024 onward in
  NS-IBTS and BITS). It occurs under both DataType P (293 groups, where it is the intended
  encoding) and DataType R (128 groups — see [Current Issues](#sec-issues)).

  `TotalNumber` and `SpeciesCategoryWeight` also vary by stratum (104 of the ~2.1 million
  valid groups, 0.005%) as a direct
  consequence. The safe formula for the haul-level total is always
  `n_haul = sum(NumberAtLength × SubsamplingFactor)` applied at the individual row level —
  this is correct whether or not size-stratification is present.

*The R code used to derive these statistics is in [Appendix 2](#sec-appendix2-constancy).*

##### Catch totals {#hl-catch-totals}

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `TotalNumber` | `TotalNo` | dbl | — | Total catch per group; semantics depend on DataType (see [Current Issues](#sec-issues)) |
| `SpeciesCategoryWeight` | `CatCatchWgt` | dbl | g | Total catch weight per group; also DataType-dependent |

`TotalNumber` and `SpeciesCategoryWeight` are repeated identically across every length row for
a given `.id` × `ValidAphiaID` × `SpeciesSex` × `DevelopmentStage` × `SpeciesCategory` group
in most records. **Do not sum `TotalNumber` across length rows** — a species with 30 length
bins produces 30 identical copies of the same total. When size-stratified subsampling is used,
`TotalNumber` differs between strata and reflects the stratum total, not the species-haul total.

**Worked example — size-stratified subsampling (NS-IBTS 2025, DK vessel, whiting
AphiaID 126437):**

| Stratum | Length range | `SubsamplingFactor` | Rows measured | `TotalNumber` (stratum) |
|---------|-------------|---------------------|---------------|------------------------|
| Small fish | 100–160 mm | 5.9 | 61 | ≈ 360 |
| Large fish | 210–420 mm | 1.0 | 168 | 168 |

Correct haul total: **360 + 168 = 528 fish**. Direct use of either stratum `TotalNumber` is
wrong.

The correct approach for all DataTypes R, S, and P:

```r
hl |>
  filter(SpeciesValidity == "1") |>
  left_join(hh |> select(.id, DataType, HaulDuration), by = ".id") |>
  group_by(.id, ValidAphiaID, SpeciesSex, DevelopmentStage, SpeciesCategory) |>
  summarise(n_haul = sum(NumberAtLength * SubsamplingFactor), .groups = "drop")
```

##### Length frequency {#hl-length-freq}

One row per length class within a species × sex × development stage × category group.

| Field | Old name | Type | Description |
|-------|----------|------|-------------|
| `LengthClass` | `LngtClass` | int | Length class value; unit determined by `LengthCode` |
| `LengthCode` | `LngtCode` | chr | Length measurement unit — [codes](#vocab-lngtcode) |
| `LengthType` | `LenMeasType` | chr | Measurement convention: total length, fork length, mantle length, etc. — [codes](#vocab-lenmeastype) |
| `NumberAtLength` | `HLNoAtLngt` | dbl | Count in the measured sub-sample at this length class; semantics depend on DataType |
| `SpeciesSex` | `Sex` | chr | Population-level sex; visual inspection of bulk catch. {obus} maps to `sex` (lowercase) — [codes](#vocab-sex), [Appendix 2](#sec-appendix2-deviations) |
| `DevelopmentStage` | `DevStage` | chr | Non-invasive maturity or developmental stage observation |

> **Note (A2):** `LengthType` is particularly relevant for species where the standard length
> convention differs by country or era. `DevelopmentStage` prevalence and the conditions under
> which it is populated require confirmation (action point A2, 1 June 2026).

`SpeciesSex` and `DevelopmentStage` are grouping dimensions within the length frequency —
they define distinct sub-populations measured separately. Both must be included in any
grouping key before aggregating across length classes.

**Naming note:** The parquet file retains the legacy name `IndividualSex` for the
population-level sex field in HL. This is a misnomer: in HL the sex is assessed visually on
the bulk catch, not by individual dissection. The proposed standard name `SpeciesSex` is used
throughout this document to distinguish it from `IndividualSex` in CA, which is
dissection-based.

---

#### CA — Catch-at-age {#sec-ca}

CA contains biological measurements on fish selected for ageing. One row per measured
individual (most rows) or per length class for pooled area-level samples.

CA shares the same eight haul join fields as HH and HL. After joining on `.id`, the haul-level
fields are redundant.

| Field | Old name | Type | Unit | Description |
|-------|----------|------|------|-------------|
| `ValidAphiaID` | `Valid_Aphia` | int | — | Resolved WoRMS AphiaID |
| `AreaCode` | `AreaCode` | chr | — | Aggregation area code; check `AreaType` before joining to HH |
| `AreaType` | `AreaType` | chr | — | `"H"` = haul-level (safe to join); other values = area-pooled — [codes](#vocab-areatype) |
| `FishID` | `FishID` | chr | — | Individual fish identifier within haul |
| `IndividualSex` | `Sex` | chr | — | Sex by dissection; distinct from `SpeciesSex` in HL. {obus} maps to `sex` (lowercase) — [codes](#vocab-sex), [Appendix 2](#sec-appendix2-deviations) |
| `LengthClass` | `LngtClass` | int | see `LengthCode` | Measured length of individual fish |
| `LengthCode` | `LngtCode` | chr | — | Length measurement unit — [codes](#vocab-lngtcode) |
| `LengthType` | — | chr | — | Measurement convention: total length, fork length, etc. — [codes](#vocab-lenmeastype) |
| `IndividualAge` | `AgeRings`^[Three-way mismatch: web service says `new = IndividualAge`, `old = AgeRings`; but `icesDatras::getDATRAS()` returns this column as `Age`. See [Appendix 2](#sec-appendix2-deviations).] | int | Years | Otolith-based age estimate |
| `AgePlusGroup` | `PlusGr` | chr | — | Plus-group flag — [codes](#vocab-plusgr) |
| `AgeSource` | `AgeSource` | chr | — | Tissue type used for ageing — [codes](#vocab-agesource) |
| `AgePreparationMethod` | `AgePrepMet` | chr | — | Preparation method — [codes](#vocab-ageprepmet) |
| `OtolithGrading` | `OtGrading` | chr | — | Reader quality score — [codes](#vocab-otgrading) |
| `IndividualWeight` | `IndWgt` | dbl | g | Individual body weight |
| `LiverWeight` | `LiverWeight` | dbl | g | Liver weight when sampled |
| `IndividualMaturity` | `Maturity` | chr | — | Maturity stage code — [codes](#vocab-maturity) |
| `MaturityScale` | `MaturityScale` | chr | — | Maturity scale used (M6 or SMSF from 2021+) — [codes](#vocab-maturityscale) |
| `NumberAtLength` | `CANoAtLngt` | dbl | — | Count at this length/age class (pooled rows only) |
| `SpeciesCodeType` | `SpecCodeType` | chr | — | Species code system used — [codes](#vocab-speccodeype) |

> **Important:** When `AreaType != "H"`, age samples are pooled across multiple hauls within
> an area. Joining these rows to a specific HH record is ambiguous. Always check `AreaType`
> before assuming a 1:1 join to HH.

---

#### CPUE and Derived Products {#sec-cpue-products}

ICES produces several derived products from the exchange tables, accessible without
downloading raw data:

| Product | Description |
|---------|-------------|
| CPUE per length (CPUEL) | Official ICES CPUE-by-length product |
| CPUE per age (CPUEA) | Age-disaggregated CPUE |
| Survey indices (IDX) | Age-based survey indices |
| Catch weight (CW) | Catch weight by species |
| Flex file (FL) | Survey-specific additional fields |
| Litter assessment (LT) | Marine litter per haul |

The official CPUEA and IDX products embed survey- and species-specific decisions (ALK
construction, plus-group handling, stratum weighting) agreed by ICES expert groups. These
decisions are not encoded in the exchange tables and cannot be replicated from exchange data
alone. WP2 does not aim to replicate these products; see [D2.3](D2.3_imbus_product_specification.qmd#cpue-pathways)
for the IMBUS exploratory CPUE-by-age approach.

---

### Meta-data Descriptions {#sec-metadata}

ICES provides field-level metadata through the DATRAS Technical Reference, covering names,
types, units, and valid values for all exchange columns. Survey-level metadata (gear
specifications, time-series breaks, QC history) are held by contributing national institutes
and are not centrally available in machine-readable form; per-survey fact sheets are a
separate deliverable outside D2.2.

#### Surveys covered

The specification applies to all ICES surveys available in DATRAS with valid hauls
(`HaulValidity == "V"`), excluding test surveys. Generation code is in
[Appendix 2](#sec-appendix2-surveys).

| Survey | First year | Last year | Valid hauls | Species (AphiaIDs) |
|--------|-----------|----------|------------|-------------------|
| NS-IBTS | 1965 | 2026 | 36,566 | 837 |
| BTS | 1985 | 2025 | 19,887 | 1,012 |
| BITS | 1991 | 2026 | 18,659 | 193 |
| DYFS | 1985 | 2025 | 17,943 | 443 |
| Can-Mar | 1970 | 2021 | 11,372 | 408 |
| NSSS | 2008 | 2025 | 5,134 | 8 |
| SP-NORTH | 1990 | 2025 | 4,455 | 404 |
| EVHOE | 1997 | 2025 | 4,001 | 304 |
| IE-IGFS | 2003 | 2025 | 3,646 | 238 |
| FR-CGFS | 1988 | 2025 | 3,161 | 280 |
| SWC-IBTS | 1985 | 2010 | 2,702 | 174 |
| SNS | 1985 | 2025 | 2,138 | 230 |
| SP-PORC | 2001 | 2025 | 2,007 | 229 |
| NIGFS | 2005 | 2025 | 1,923 | 114 |
| SCOWCGFS | 2011 | 2026 | 1,790 | 168 |
| SP-ARSA | 1996 | 2025 | 1,732 | 246 |
| PT-IBTS | 2002 | 2023 | 1,551 | 392 |
| BTS-VIII | 2007 | 2025 | 1,437 | 113 |
| BTS-GSA17 | 2008 | 2025 | 1,128 | 8 |
| IE-IAMS | 2016 | 2025 | 1,105 | 192 |
| SCOROC | 2011 | 2025 | 636 | 90 |
| NL-BSAS | 2019 | 2025 | 408 | 2 |
| ROCKALL | 1999 | 2009 | 364 | 68 |
| FR-WCGFS | 2018 | 2025 | 361 | 97 |
| NS-IDPS | 2009 | 2016 | 250 | 21 |
| SE-SOUND | 2011 | 2025 | 155 | 50 |
| IS-IDPS | 2009 | 2009 | 89 | 67 |
| DWS | 2006 | 2009 | 71 | 173 |
| CODS-Q4 | 2024 | 2025 | 42 | 42 |

NS-IBTS is the largest survey (36,566 valid hauls from 1965); CODS-Q4 is the smallest (42
hauls from 2024). BTS records the most distinct species (1,012 AphiaIDs). `SWC-IBTS` ended
in 2010 and is superseded by `SCOWCGFS`.

#### Field and vocabulary reference {#sec-lookup}

{obus} exposes the DATRAS field-level and vocabulary metadata as two data objects:

**`dr_lookup_fields`** is the complete field dictionary for all DATRAS exchange and derived
product tables (HH, HL, CA, FL, LT, CPUEL, CPUEA, IDX). For each field it records the
standard name (`new`), the legacy name (`old`), the data type (`chr`/`int`/`dbl`), and a
plain-English description. The primary source is the ICES DATRAS web service endpoint
`getDatrasFieldList`
(`https://datras.ices.dk/WebServices/DATRASWebService.asmx/getDatrasFieldList`), which
covers the formal upload schema for HH, HL, CA, and FA. This is supplemented by
hand-curated entries for tables and fields not covered by the upload schema: LT and FL
(returned by `icesDatras` helper functions), the derived products CPUEL, CPUEA, and IDX,
and the `Valid_Aphia` / `aphia` mapping — a field added by the ICES datacenter during
processing that is absent from the formal upload field list. The script that builds
`dr_lookup_fields` is `data-raw/DATASET_lookup_fields.R` in the {obus} repository.

**`dr_lookup_vocabulary`** provides the controlled vocabularies for categorical DATRAS
fields — i.e. the valid codes and their descriptions for fields such as `HaulValidity`,
`DataType`, `LengthCode`, `Gear`, and `Country`. The source is the ICES Vocabulary service,
accessed via the `icesVocab` R package (`getCodeTypeList()` + `getCodeList()`). The full
ICES vocabulary is filtered to the subset of code types whose variable names correspond to
fields in `dr_lookup_fields`, giving one row per valid code value. The building script is
`data-raw/DATASET_vocabulary.R` in the {obus} repository.

```r
library(obus)

# Field dictionary: names, types, descriptions for all DATRAS tables
dr_lookup_fields

# Controlled vocabularies: valid codes for categorical fields
dr_lookup_vocabulary
```

These two objects together constitute the machine-readable metadata layer underlying the
field tables in §2. Both are regenerated by sourcing the respective `data-raw/` scripts
against live ICES web services; the versions shipped with {obus} reflect the state of
those services at the time of the last package build.

The build scripts apply corrections and additions to the raw web service output; see
[Appendix 2 — Build script deviations](#sec-appendix2) for the full account.

##### Controlled vocabulary tables {#sec-vocab}

The tables below list the valid codes for categorical fields. They are generated directly
from `dr_lookup_vocabulary` and are linked from the relevant rows in the field tables
above. Fields with large open-ended vocabularies (e.g. `Gear`, `Country`,
`StatisticalRectangle`) are excluded.

###### `HaulValidity` (HaulVal) {#vocab-haulval}

```{r}
dr_lookup_vocabulary |>
  filter(old == "HaulVal") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `DataType` {#vocab-datatype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "DataType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `ThermoCline` {#vocab-thermocline}

```{r}
dr_lookup_vocabulary |>
  filter(old == "ThermoCline") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `PelagicSamplingType` (PelSampType) {#vocab-pelsamtype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "PelSampType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `DoorType` {#vocab-doortype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "DoorType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `GearExceptions` (GearEx) {#vocab-gearex}

```{r}
dr_lookup_vocabulary |>
  filter(old == "GearEx") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `Rigging` {#vocab-rigging}

```{r}
dr_lookup_vocabulary |>
  filter(old == "Rigging") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `SpeciesValidity` (SpecVal) {#vocab-specval}

```{r}
dr_lookup_vocabulary |>
  filter(old == "SpecVal") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `SpeciesCodeType` (SpecCodeType) {#vocab-speccodeype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "SpecCodeType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `LengthCode` (LngtCode) {#vocab-lngtcode}

Used in both HL and CA.

```{r}
dr_lookup_vocabulary |>
  filter(old == "LngtCode") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `LengthType` (LenMeasType) {#vocab-lenmeastype}

Used in both HL and CA.

```{r}
dr_lookup_vocabulary |>
  filter(old == "LenMeasType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `Sex` — `SpeciesSex` (HL) / `IndividualSex` (CA) {#vocab-sex}

```{r}
dr_lookup_vocabulary |>
  filter(old == "Sex") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `AgePlusGroup` (PlusGr) {#vocab-plusgr}

```{r}
dr_lookup_vocabulary |>
  filter(old == "PlusGr") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `AgeSource` {#vocab-agesource}

```{r}
dr_lookup_vocabulary |>
  filter(old == "AgeSource") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `AgePreparationMethod` (AgePrepMet) {#vocab-ageprepmet}

```{r}
dr_lookup_vocabulary |>
  filter(old == "AgePrepMet") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `OtolithGrading` (OtGrading) {#vocab-otgrading}

```{r}
dr_lookup_vocabulary |>
  filter(old == "OtGrading") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `IndividualMaturity` (Maturity) {#vocab-maturity}

```{r}
dr_lookup_vocabulary |>
  filter(old == "Maturity") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `MaturityScale` {#vocab-maturityscale}

```{r}
dr_lookup_vocabulary |>
  filter(old == "MaturityScale") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

###### `AreaType` {#vocab-areatype}

```{r}
dr_lookup_vocabulary |>
  filter(old == "AreaType") |>
  distinct(key, description) |>
  arrange(key) |>
  kable(col.names = c("Code", "Description"))
```

---

### Current Issues with DATRAS Data Formats and Meta-data Descriptions {#sec-issues}

#### Column naming convention

Two naming conventions exist in DATRAS, reflecting a change introduced by ICES around 2020:

| Convention | Source | Example fields |
|------------|--------|----------------|
| **Legacy** | `getDATRAS()` (XML API); pre-~2020 DATRAS downloads | `Ship`, `HaulNo`, `StNo`, `LngtCode`, `HLNoAtLngt`, `SubFactor`, `TotalNo`, `HaulVal`, `HaulDur` |
| **Standard** | Parquet files; IMBUS product; current ICES convention | `Platform`, `HaulNumber`, `StationName`, `LengthCode`, `NumberAtLength`, `SubsamplingFactor`, `TotalNumber`, `HaulValidity`, `HaulDuration` |

The co-existence of both conventions means that code written against one naming scheme fails
silently or with cryptic errors against the other. All field tables in §2.1 list both names.

#### The DataType field {#sec-datatype}

`DataType` is set at the haul level (HH) and governs how `NumberAtLength`, `TotalNumber`, and
`SpeciesCategoryWeight` in HL must be interpreted. It is the single most common source of
silent errors when working with exchange data, because the numeric meaning of HL fields depends
on a field in a different table.

| DataType | Sampling protocol | `SubsamplingFactor` | `NumberAtLength` | `n_haul` formula |
|----------|-------------------|---------------------|------------------|-----------------|
| `R` | Raw catch; some species sub-sampled | ≥ 1 | Count in sub-sample | `NumberAtLength × SubsamplingFactor` |
| `S` | Unsorted bulk sub-sample | > 1 | Count in sub-sample | `NumberAtLength × SubsamplingFactor` |
| `P` | Size-stratified (pseudocategory) | ≥ 1 per stratum | Count in sub-sample for stratum | `NumberAtLength × SubsamplingFactor` |
| `C` | Pre-raised to 1-hour haul | 1 | Numbers per hour (not a raw count) | `NumberAtLength × HaulDuration / 60` |

For DataTypes R, S, P: `n_hour = n_haul / HaulDuration × 60`  
For DataType C: `n_hour = NumberAtLength` (already per hour)

`TotalNumber` and `SpeciesCategoryWeight` are similarly DataType-dependent:

- **R/S/P:** total per-haul count and weight
- **C:** total per-hour count and weight (multiply by `HaulDuration / 60` to recover per-haul
  values)

**Size-stratified subsampling under DataType R.** The DATRAS schema intends DataType P for
hauls processed with size-stratified subsampling. In practice, GB-SCT vessels (293 groups
across NS-IBTS, SCOWCGFS, SCOROC) and DK vessels (82 groups from 2024 onward) submit
size-stratified hauls as DataType R. These are not data errors — the row-level formula
`NumberAtLength × SubsamplingFactor` remains correct. The risk arises only from direct use
of `TotalNumber`, which reflects the stratum total, not the species-haul total, in these
records.

*The R code used to derive these statistics is in [Appendix 2](#sec-appendix2-datatype).*

`DataType` must be joined from HH before any HL arithmetic:

```r
hl |> left_join(hh |> select(.id, DataType, HaulDuration), by = ".id")
```

#### Cross-table ambiguities

Several fields share names or concepts across tables but differ in meaning.

**Across HL and CA:**

| Issue | HL | CA |
|-------|----|----|
| Sex field | `SpeciesSex` (`IndividualSex` in parquet) — visual, population-level | `IndividualSex` — determined by dissection |
| `NumberAtLength` | Sub-sample count per length bin; multiply by `SubsamplingFactor` for population estimate | Count of aged fish at that length class; used as raising key for ALK |
| `LengthClass` | Lower boundary of a length frequency bin | Measured length of an individual fish |

**Gear fields duplicated across HH, HL, and CA:**

`Gear`, `GearExceptions`, `SweepLength`, and `DoorType` appear in all three tables with
identical values. This is a submission artefact. HH is the sole source of truth; the HL and
CA copies are redundant after joining on `.id`.

#### Other known data issues

- **`SubsamplingFactor = NA`:** Approximately 0.015% of HL records have missing
  `SubsamplingFactor` under DataType R (Norwegian and Dutch vessels in NS-IBTS). This
  propagates as `NA` in computed `n_haul` and `n_hour`.
- **`TotalNumber` rounding for DataType C:** `TotalNumber` is a back-converted CPUE count;
  arithmetic checks reveal rounding differences of 1–20 fish (~1% of DataType C groups). This
  is an artefact of the ICES submission process and is expected.
- **CA `AreaCode`/`AreaType`:** CA rows can represent area-pooled age samples, not individual
  haul observations. Check `AreaType` before joining CA to HH.
- **Area standardisation:** Mapping from haul positions to C-squares and ICES divisions is
  not provided in the current DATRAS exchange format; this is addressed in D2.3.

---

## Discussion and Conclusions {#sec-discussion}

*[To be completed.]*

---

## Further Work {#sec-further-work}

The following items are open for D2.2 as of the date of this document:

1. **CA section.** Field-by-field treatment of CA comparable to the HH and HL sections is
   needed; the current CA section is a stub.

2. **Per-survey meta-data fact sheets.** Coverage, gear history, time-series breaks, and QC
   caveats at the survey level (user requirement UR-Q1) are a separate deliverable not
   covered by D2.2.

Items relating to the proposed parquet product, {obus}, parquet hosting, D5.3 integration,
and area standardisation are tracked in
**[D2.3](D2.3_imbus_product_specification.qmd#sec-further-work)**.

---

## References {#sec-references}

*[To be completed.]*

- ICES DATRAS Technical Reference (available from the ICES Datacenter)
- IMBUS D2.1 — User requirements (April 2026)
- WoRMS Editorial Board (2024). World Register of Marine Species. <https://www.marinespecies.org>

---

## Annex 1 — Related Documents and Resources {#sec-annex}

The following companion resources are part of the IMBUS WP2 output but are not reproduced
in the body of this deliverable:

- **[D2.3 — IMBUS Integrated Data Product Specification](D2.3_imbus_product_specification.qmd)**
  — Proposed standardised parquet product; data tiers; {obus} access; CPUE pathways.
- **[DATRAS Field Reference](datras_field_reference.qmd)** — Complete field reference tables
  for all HH, HL, CA, and LT columns; length-unit and species-coding rules; data quality
  checks.
- **`docs_external/DATRAS_documentation/DATRAS_FAQ.qmd`** — Restructured DATRAS FAQ with
  updated field names.
- **`docs_external/DATRAS_Technical_Reference.md`** — Full DATRAS field and unit reference
  (ICES).
- **`background/IMBUS_proposal_summary.qmd`** — IMBUS project context and WP2 objectives.
- **`meetings/20260601_WP2-WP5.qmd`** — Minutes of the WP2-WP5 alignment meeting,
  1 June 2026.

---

## Appendix 2 — Redundant Sections {#sec-appendix2}

Sections moved here from the main document to reduce length. They remain part of D2.2 but
are not required reading for reviewers of the field format documentation.

### Scope {#sec-scope}

**In scope:**

- The three DATRAS exchange record types: **HH**, **HL**, and **CA**.
- Field-level metadata: names, types, units, descriptions, and known interpretation pitfalls.
- The field structure of HL and the conceptual groups within it.
- The `DataType` field and its arithmetic implications for HL fields.
- Species resolution: the WoRMS AphiaID system and historical code types.
- Survey coverage: year ranges, valid haul counts, and distinct species per survey.
- Known issues with the current DATRAS format (naming, DataType, cross-table ambiguities).
- Machine-readable field metadata via `dr_lookup_fields` and `dr_lookup_vocabulary` ({obus}).

**Out of scope for D2.2** (covered in D2.3):

- The proposed IMBUS standardised parquet product and data tiers.
- Parquet-based distribution and the {obus} access layer.
- The unique haul identifier (`.id`) construction.
- CPUE calculation pathways and intermediate data products.
- Area standardisation.
- **Official CPUEA and survey indices:** ICES Datacenter products; accessible via
  `dr_get("CPUEA", ...)` and `dr_get("IDX", ...)`.
- **Deployed data service:** Deployment to the ICES open access portal is the subject of D2.4.
- **Per-survey metadata fact sheets:** Addressed separately under user requirement UR-Q1.

### Build script deviations from upstream sources {#sec-appendix2-deviations}

The `data-raw/` scripts in {obus} apply corrections and additions to the raw ICES web
service output when building `dr_lookup_fields` and `dr_lookup_vocabulary`.

**`dr_lookup_fields` — deviations from the ICES web service**

*Format normalisation.* The web service returns format codes in ICES-internal notation;
the script maps these to standard R type names: `"char"` → `"chr"`, `"decimal"` /
`"float"` → `"dbl"`. Additionally, `Year` is corrected from `"char"` to `"int"` (the
web service value is incorrect), and CA `NumberAtLength` is forced to `"dbl"` to avoid
type ambiguity with the identically-named HL field.

*`IndividualAge` / `AgeRings` / `Age` — three-way naming mismatch.* The web service
field list returns `new = "IndividualAge"` and `old = "AgeRings"` for the CA age field.
However, `icesDatras::getDATRAS(source = "xml")` returns this column as simply `Age` —
matching neither the web service new name nor the web service old name. The build script
corrects both to `new = "Age"`, `old = "Age"` to align with the actual `getDATRAS()`
column. The tables in this document follow the web service values (`IndividualAge` /
`AgeRings`) as these reflect the formal ICES field definition; users working with
`getDATRAS()` output will see `Age` instead.

*Sex field naming deviation (`sex`, lowercase).* The web service returns
`new = "SpeciesSex"`, `old = "Sex"` for HL and `new = "IndividualSex"`, `old = "Sex"`
for CA. The build script deliberately maps both to `new = "sex"` (lowercase) as a
temporary canonical name, treating the two as the same concept until the ICES Datacenter
resolves the upstream inconsistency. D2.2 retains the ICES field names (`SpeciesSex` in
HL, `IndividualSex` in CA).

*`Valid_Aphia` / `aphia` and `DateofCalculation` added.* These fields are not part of
the formal DATRAS upload schema and do not appear in `getDatrasFieldList`. They are added
by the ICES datacenter during processing. The build script adds them manually:
`new = "aphia"`, `old = "Valid_Aphia"` for HL and CA; `DateofCalculation` for HH, HL,
and CA. D2.2 and D2.3 propose `ValidAphiaID` as the canonical name for `Valid_Aphia`.

*Hand-curated tables.* The ICES web service covers only the upload schema (HH, HL, CA,
FA). The following tables are entirely hand-curated:

| Table | Reason |
|-------|--------|
| LT | Web service LT entries use new-style names that do not match `getLTassessment()` actual output (`Ship`, `StNo`, `HaulNo`) |
| FL | Columns returned by `icesDatras::getFlexFile()`; no web service coverage |
| CPUEL, CPUEA, IDX | Derived products; no web service coverage |

**`dr_lookup_vocabulary` — deviations from the ICES Vocabulary Service**

The ICES Vocabulary Service uses type code prefixes (`TS_`, `AC_`) that do not appear in
DATRAS field names. The build script strips these prefixes (e.g. `TS_HaulVal` →
`HaulVal`) to allow matching against `dr_lookup_fields$old`. `Month`, `Quarter`, and
`Year` are excluded as their vocabulary entries are not useful for categorical lookups.

### Verification code {#sec-appendix2-code}

#### HL sampling protocol constancy checks {#sec-appendix2-constancy}

The tables below are generated from the parquet files and reproduce the statistics cited
in [Sampling protocol](#hl-sampling-protocol). Column names are legacy (parquet) names.

The code below reproduces the statistics cited in [Sampling protocol](#hl-sampling-protocol).
Results are pre-computed and cited in the main text; this code is provided for reproducibility.
Column names are legacy (parquet) names. Run interactively — these queries scan the full HL
parquet and are too slow to execute during document rendering.

```r
library(obus)
library(dplyr)

hl  <- dr_con("HL")
hh  <- dr_con("HH")
grp <- c(".id", "Valid_Aphia", "IndividualSex", "DevelopmentStage", "SpeciesCategory")

# Total groups (all SpeciesValidity): 2,588,836
hl |> group_by(across(all_of(grp))) |> summarise(.groups = "drop") |> count() |> collect()

# Total valid groups (SpeciesValidity == "1"): 2,100,517
hl |> filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |> summarise(.groups = "drop") |> count() |> collect()

# SpeciesValidity: groups with > 1 distinct value (2,786 of 2,588,836 = 0.1%)
specval_v <- hl |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(SpeciesValidity), .groups = "drop") |>
  filter(n_vals > 1)

specval_v |>
  left_join(hl |> distinct(.id, Survey), by = ".id") |>
  count(Survey, sort = TRUE) |> collect()            # BTS 1,289; Can-Mar 1,272

hl |>
  semi_join(specval_v, by = grp) |>
  group_by(across(all_of(grp))) |>
  summarise(combo = paste(sort(unique(SpeciesValidity)), collapse = ", "),
            .groups = "drop") |>
  count(combo, sort = TRUE) |> collect()             # {1, 5}: 1,281; {4, 7}: 580

# SubsamplingFactor: groups with > 1 distinct value (389 of 2,100,517 = 0.02%)
sf_v <- hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_vals > 1)

sf_v |>
  left_join(hl |> distinct(.id, Survey, Country), by = ".id") |>
  count(Country, Survey, sort = TRUE) |> collect()   # GB-SCT 293; DK 82

sf_v |>
  left_join(hh |> distinct(.id, DataType), by = ".id") |>
  count(DataType) |> collect()                       # P: 293; R: 128 (? - 389 total)

# TotalNumber: groups with > 1 distinct value (104 of ~2.1 million valid groups = 0.005%)
tn_v <- hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_vals = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_vals > 1)

tn_v |> count() |> collect()                        # 104

tn_v |>
  left_join(hl |> distinct(.id, Survey), by = ".id") |>
  count(Survey, sort = TRUE) |> collect()
```

#### DataType R size-stratified hauls {#sec-appendix2-datatype}

Code used to derive the size-stratified DataType R statistics reported in [The DataType
field](#sec-datatype).

```r
library(obus)
library(dplyr)

hl  <- dr_con("HL")
hh  <- dr_con("HH")
grp <- c(".id", "Valid_Aphia", "IndividualSex", "DevelopmentStage", "SpeciesCategory")

sf_violations_R <- hl |>
  filter(SpeciesValidity == "1") |>
  group_by(across(all_of(grp))) |>
  summarise(n_sf = n_distinct(SubsamplingFactor), .groups = "drop") |>
  filter(n_sf > 1) |>
  left_join(hh |> distinct(.id, DataType, Country), by = ".id") |>
  filter(DataType == "R")

sf_violations_R |> count()                          # 128 groups
sf_violations_R |> count(Country, sort = TRUE)      # GB-SCT: 97; DK: 31 (illustrative)
```

#### Surveys table generation {#sec-appendix2-surveys}

Code used to generate the surveys table in [Surveys covered](#sec-metadata).

```r
library(obus)
library(dplyr)

hh <- dr_con("HH") |> collect()
hl <- dr_con("HL")

survey_stats <- hh |>
  filter(HaulValidity == "V", Survey != "NS-IBTS_UNIFtest") |>
  group_by(Survey) |>
  summarise(year_min = min(Year), year_max = max(Year), n_hauls = n(), .groups = "drop") |>
  arrange(desc(n_hauls))

species_per_survey <- hl |>
  filter(!is.na(Valid_Aphia), Valid_Aphia > 0) |>
  group_by(Survey) |>
  summarise(n_species = n_distinct(Valid_Aphia), .groups = "drop") |>
  collect()

survey_stats |> left_join(species_per_survey, by = "Survey")
```
